---
title: "Success of Leader Assassination as a Natural Experiment"
output:
  html_document:
    df_print: paged
---

One longstanding debate in the study of international relations
concerns the question of whether individual political leaders can make
a difference.  Some emphasize that leaders with different ideologies
and personalities can significantly affect the course of a nation.
Others argue that political leaders are severely constrained by
historical and institutional forces.  Did individuals like Hitler,
Mao, Roosevelt, and Churchill make a big difference?  The difficulty
of empirically testing these arguments stems from the fact that the
change of leadership is not random and there are many confounding
factors to be adjusted for.

In this exercise, we consider a *natural experiment* in which the
success or failure of assassination attempts is assumed to be
essentially random.

This exercise is based on:
Jones, Benjamin F, and Benjamin A Olken. 2009. “[Hit or Miss? 
 The Effect of Assassinations on Institutions and 
 War.](http://dx.doi.org/10.1257/mac.1.2.55)” 
 *American Economic Journal: Macroeconomics* 1(2): 55–87. 

Each observation of the CSV data set
`leaders.csv` contains information about an assassination
attempt.  The variables are:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `country`            The name of the country
 
 `year`               Year of assassination
 
 `leadername`         Name of leader who was targeted
 
 `age`                Age of the targeted leader
 
 `politybefore`       Average polity score during the 3 year period prior to the attempt
 
 `polityafter`        Average polity score during the 3 year period after the attempt
 
 `civilwarbefore`     1 if country is in civil war during the 3 year period prior to the attempt, or 0 
 
 `civilwarafter`      1 if country is in civil war during the 3 year period after the attempt, or 0
 
 `interwarbefore`     1 if country is in international war during the 3 year period prior to the attempt, or 0
 
 `interwarafter`      1 if country is in international war during the 3 year period after the attempt, or 0
 
 `result`             Result of the assassination attempt, one of 10 categories described below
--------------------------------------------------------------------------------

The `polity` variable represents the so-called *polity score*
from the Polity Project.  The Polity Project systematically documents
and quantifies the regime types of all countries in the world from
1800.  The polity score is a 21-point scale ranging from -10
(hereditary monarchy) to 10 (consolidated democracy).  

The `result` variable is a 10 category factor variable describing
the result of each assassination attempt.

## Question 1

How many assassination attempts are recorded in the data?  How
many countries experience at least one leader assassination attempt?
(The `unique()` function, which returns a set of unique values
from the input, may be useful here).  What is the average
age of a leader across all assassination attempts in the data?

## Answer 1

```{r}
leaders <- read.csv("data/leaders.csv")
## dimension one of assassinations is the number of rows
dim(leaders)
## number of countries in the data
head(unique(leaders$country))
## average age
mean(leaders$age)
```

The number of assassination attempts in the data is equal to the
number of observations, `r nrow(leaders)`. There are
`r length(unique(leaders$country))` countries in the dataset that
have experienced at least one assassination attempt. The average
age of a leader at the time of their assassination attempt across 
all countries and all years is equal to 
`r round(mean(leaders$age), 2)`.

## Question 2

Create a new binary variable named `success` that is equal
  to 1 if a leader dies from the attack and to 0 if the leader
  survives.  Store this new variable as part of the original data
  frame.  What is the overall success rate of leader assassination?
  Does the result speak to the validity of the assumption that the
  success of assassination attempts is randomly determined? 
  **Hint:** Use the logical operator `|` to help code this new 
  binary variable. The logical operator `|` can be used to add 
  optional criteria to a logical test, e.g., `variable.one == 1 | variable.two == 2`
  is a test that can only be "passed" with a value of `TRUE` if 
  either $variable.one = 1$ *or* $variable.two = 2$.

## Answer 2

```{r}
unique(leaders$result) 
## create variable, 1 if died "from" attack, 0 otherwise
leaders$success <- 
    ifelse(leaders$result == "dies between a day and a week" | 
             leaders$result == "dies between a week and a month" |
               leaders$result == "dies within a day after the attack" | 
             leaders$result == "dies, timing unknown", 1, 0)
## rate of success
mean(leaders$success)
```

Assassination attempts are successful
`r round(mean(leaders$success), 3)*100` percent of the time. It is
unclear whether this can validate the assumption that attempts are
randomly determined. It at least validates that not all attempts lead
to the death of the leader.  However, we do not know if other
systematic factors, such as skill of the perpetrator, health of the
leader, or security conditions, might be at work in determining
success and failure of the attempts.

## Question 3

Investigate whether the average polity score over 3 years prior
  to an assassination attempt differs on average between successful
  and failed attempts.  Also, examine whether there is any difference
  in the age of targeted leaders between successful and failed
  attempts.  Briefly interpret the results in light of the validity of
  the aforementioned assumption.

## Answer 3

```{r}
## avg polity score for successful/unsuccessful before
mean(leaders$politybefore[leaders$success == 1])
mean(leaders$politybefore[leaders$success == 0])
## avg age for successful/unsuccessful
mean(leaders$age[leaders$success == 1])
mean(leaders$age[leaders$success == 0])
```

Prior to the attempt, countries with unsuccessful assassinations were
somewhat less democratic on average than countries with successful
attempts. The age of the leaders was relatively similar, although
countries whose leaders died from the attack had slightly older
leaders. These two findings suggest that there may be some systematic
differences in the types of countries and types of leaders who are
more susceptible to successful assassination attempts, though the
differences may not be large.

## Question 4

Repeat the same analysis as in the previous question, but this
  time using the country's experience of civil and international war.
  Create a new binary variable in the data frame called
  `warbefore`.  Code the variable such that it is equal to 1 if
  a country is in either civil or international war during the 3 years
  prior to an assassination attempt.  Provide a brief interpretation
  of the result.

## Answer 4

```{r}
## create `warbefore' variable
leaders$warbefore <- ifelse(leaders$interwarbefore == 1 |
                                leaders$civilwarbefore == 1, 1, 0)
## proportion war before successful/unsuccessful
mean(leaders$warbefore[leaders$success == 1])
mean(leaders$warbefore[leaders$success == 0])
```

Countries with successful attempts were no more or less likely to be
engaged in war prior to the assassination attempt than those with
unsuccessful attempts.  This finding is consistent with the assumption
that the success of assassination attempts is random.

## Question 5

Does successful leader assassination cause democratization?
  Does successful leader assassination lead countries to war?  Answer
  these questions by analyzing the data.  Be sure to state your
  assumptions and provide a brief interpretation of the results.

## Question 5

Does successful leader assassination cause democratization?
  Does successful leader assassination lead countries to war? 
  To try to answer these questions, we can employ a 
  *difference-in-differences* analysis. Implement this analysis by taking
  the difference in mean polity scores before and
  after the assassination attempts for countries with unsuccessful
  attempts and subtracting this difference from the difference score for
  those with successful attempts. The assumption here is that the change
  in polity scores for countries with successful attempts would be the
  same as the change for those with unsuccessful attempts had their
  assassinations not been successful. Be sure to state the validity
  of these assumptions in this context
  and provide a brief interpretation of the results.

## Answer 5

```{r}
## compare polity scores before to scores after
diff.pol.succ <- mean(leaders$polityafter[leaders$success == 1]) -
    mean(leaders$politybefore[leaders$success == 1]) # among successful
diff.pol.unsucc <-mean(leaders$polityafter[leaders$success == 0]) -
    mean(leaders$politybefore[leaders$success == 0]) # unsuccessful
## difference in differences
diff.pol.succ - diff.pol.unsucc
```

The result of differences-in-differences is small and negative,
suggesting that successful attempts do not lead to democratization.
However, the difference-in-differences assumptions may not readily 
hold in this context. Even on average, there are potentially many reasons why
countries that had successful assassination 
attempts might have had different polity score
time trends in the absence of success than their factually unsuccessful
counterparts, e.g., any number of time-varying confounders pertaining to
the intricacies and complexities of political economy, from weaker institutions to different
(unobservable) cultural norms and forces.
