---
title: "Voting in the United Nations General Assembly"
output:
  html_document:
    df_print: paged
---

Like legislators in the US Congress, the member states of the United
Nations (UN) are politically divided on many issues such as trade,
nuclear disarmament, and human rights.  During the Cold War, countries
in the UN General Assembly tended to split into two factions: one led
by the capitalist United States and the other by the communist Soviet
Union.  In this exercise, we will analyze how states' ideological
positions, as captured by their votes on UN resolutions, have changed
since the fall of communism.

This exercise is based on Michael A. Bailey, Anton Strezhnev, and 
Erik Voeten. "Estimating Dynamic State Preferences from United Nations Voting Data."
 *Journal of Conflict Resolution*, August 2015.

The data is called `unvoting.csv` and the variables are:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `CountryName`        The name of the country
 
 `idealpoint`         Its estimated ideal point
 
 `Year`               The year for which the ideal point is estimated
 
 `PctAgreeUS`         The percentage of votes that agree with the US 
                      on the same issue
 
 `PctAgreeRUSSIA`     The percentage of votes that agree with Russia/the Soviet 
                      Union on the same issue
--------------------------------------------------------------------------------

We will also make use of derivative data in files `unvotingYearAvg.csv` 
and `formersoviet.csv`.

In the analysis that follows, we measure state preferences in two
ways. Note that the data for 1964 are missing due to the absence of
roll call data. 

First, we can use the percentage of votes by each country that
coincide with votes on the same issue cast by the two major Cold War
powers: the United States and the Soviet Union. For example, if a
country voted for ten resolutions in 1992, and if its vote matched the
United States's vote on exactly six of these resolutions, the variable
`PctAgreeUS` in 1992 would equal 60 for this country. 

Second, we
can also measure state preferences in terms of numerical ideal points.  
These ideal
points capture what international relations scholars have called
countries' *liberalism* on issues such as political freedom,
democratization, and financial liberalization. The two measures are
highly correlated, with larger (more liberal) ideal points
corresponding to a higher percentage of votes that agree with the US.

## Question 1

We begin by examining how the distribution of state ideal points
   has changed since the end of communism. Plot the distribution of
   ideal points separately for 1980 and 2000 - about ten years before
   and after the fall of the Berlin Wall, respectively - using the
   function `hist()`.  Add median to each plot as a vertical line.  How
   do the two distributions differ?
   Pay attention to the degree of polarization and give a brief
   substantive interpretation of the results. **Hint:**
   To simplify the code, we can use the function `subset()` 
   to create a new dataframe for each period. `subset()` takes
   two arguments, a dataframe followed by 
   logical test, and returns a subsetted dataframe where all
   observations pass the given logical test.

## Answer 1

```{r, ans1}
## Load data
un <- read.csv("data/unvoting.csv")
## Create subsets
un1980 <- subset(un, subset = Year == 1980)
un2000 <- subset(un, subset = Year == 2000)
## Generate histograms
hist(un1980$idealpoint, xlim = c(-3, 3), xlab = "Ideal points",
    main = "1980")
abline(v = median(un1980$idealpoint))
hist(un2000$idealpoint, xlim = c(-3, 3), xlab = "Ideal points",
    main = "2000")
abline(v = median(un2000$idealpoint))
```

Perhaps surprisingly, the fall of communism did not lead to
ideological convergence across states.  Rather, in 2000, most states
clustered at around either 1 or -1.  Although the range of the
distribution moved in a more liberal direction, the median and modal
UN member-state became less liberal over time.

## Question 2

Next, examine how the number of countries voting with the US has
  changed over time. To do so, we will load and use the data in
  `unvotingYearAvg.csv`, which is a transformed version of 
  `unvoting.csv` containing yearly averages and medians 
  of all its numeric variables. Plot the average percent agreement with US across
  all counties over time.  Note that the function we use to make
  scatter plots, `plot()`, can also be made to produce line plots
  by specifying the optional argument `type` as `"l"`. Also, add 
  the average percent agreement
  with Russia as another line for comparison using the function
  `lines()`. `lines()` behaves like `abline()` and 
  takes the exact same arguments as `plot()`.  Does the US appear to
  be getting more or less isolated over time, as compared to Russia?
  Give a brief substantive interpretation
  of the results.

## Answer 2

```{r}
## Load yearly average data
un.year.avg <- read.csv("data/unvotingYearAvg.csv")
## Plot
plot(x = un.year.avg$Year, y = un.year.avg$meanPctAgreeUS, 
     ylim = c(0, 1), xlab = "Year", ylab = "Average percent agreement", 
     type = "l", col = "blue")
lines(x = un.year.avg$Year, y = un.year.avg$meanPctAgreeRUSSIA, 
     ylim = c(0, 1), xlab = "Year", ylab = "Average percent agreement", 
     type = "l", col = "red")
```

In recent years, there have been far fewer pro-US countries than pro-
Russia countries. In fact, since the end of World War II, Russia
appears to have gotten more and more popular while the US has
consistently been isolated.

## Question 3

One problem of using the percentage of votes that agree with the
  US or Russia as a measure of state preferences is that the
  ideological positions, and consequently the voting patterns, of the
  two countries might have themselves changed over time.  This makes
  it difficult to know which countries' ideological positions have
  changed.  Investigate this issue by plotting the evolution of the
  two countries' ideal points over time. Add the yearly median ideal
  point of all countries. How might the results of this analysis
  modify (or not) your interpretation of the previous analysis? **Hint:**
  The `Year` variable in `unvotingYearAvg.csv` contains all 
  unique years in the data in increasing order.

## Answer 3

```{r}
plot(x = un.year.avg$Year, y = un$idealpoint[un$CountryAbb == "USA"], 
     ylim = c(-3, 3), xlab = "Year", ylab = "Ideal Point",
     type = "l", col = "blue")
lines(x = un.year.avg$Year, y = un$idealpoint[un$CountryAbb == "RUS"], 
      col = "red")
lines(x = un.year.avg$Year, y = un.year.avg$medianIdealpoint)
```

The United States has remained more liberal than the median UN
member-state, but this distance has grown over time. On the other
hand, Russia has become more liberal than the median UN member-state,
but is much closer to the median than the US. This suggests that
Russia is becoming more popular not because countries are moving
closer to Russia's ideology, but because Russia has moved closer to
the median after the fall of the Soviet Union.

## Question 4

Let's examine how countries that were formerly part of the
  Soviet Union differ in terms of their ideology and UN voting
  compared to countries that were not part of the Soviet Union.  The
  former Soviet Union countries are: Estonia, Latvia, Lithuania,
  Belarus, Moldova, Ukraine, Armenia, Azerbaijian, Georgia,
  Kazakhstan, Kyrgyzstan, Tajikistan, Uzbekistan, and Russia. 
  (A variable containing these country names can be found in
  the file `formersoviet.csv`.) The
  `%in%` operator, which is used as `x %in% y`, may be
  useful.  This operator returns a logical vector whose element is
  `TRUE` (`FALSE`) if the corresponding element of vector
  `x` is equal to a value contained in vector `y`.  Focus
  on the most recently available UN data from 2012 and plot each
  Post-Soviet Union state's ideal point against the percentage of its
  votes that agree with the United States. Compare the post Soviet
  Union states within the same plot against the other countries using
  the function `points()`. `points()` behaves like 
  `abline()` and takes the same arguments as `plot()`.
  Briefly comment on what you observe.

## Answer 4

```{r}
## Load names
post.soviet <- read.csv("data/formersoviet.csv")$formerSovCountryName
## An indicator variable for post Soviet Union states
un$post.soviet <- ifelse(un$CountryName %in% post.soviet, 1, 0)
## Plot
un12 <- subset(un, subset = (Year == 2012))
plot(x = un12$idealpoint[un12$post.soviet == 1], 
     y = un12$PctAgreeUS[un12$post.soviet == 1], pch = 16, 
     xlim = c(-3, 3), ylim = c(0, 1), col = "red", 
     xlab = "Ideal Points", 
     ylab = "Proportion of Votes Agreeing with US")
points(x = un12$idealpoint[un12$post.soviet == 0], 
       y = un12$PctAgreeUS[un12$post.soviet == 0], pch = 17, 
       col = "blue")
```

Interestingly, the ideal points of Post-Soviet countries are located
in the center of the distribution of all countries.  The distribution
of non post-Soviet countries is much more spread. Many Central Asian
countries remain non-liberal.  However, countries such as Estonia,
Latvia, and Lithuania are among the most liberal countries in 2012,
similar to countries such as Belgium, Denmark, and the Netherlands.

## Question 5

We have just seen that while some post-Soviet countries have
  retained non-liberal ideologies, other post-Soviet countries were
  much more liberal in 2012. Let's examine how the median ideal points
  of Soviet/post-Soviet countries and all other countries has varied
  over all the years in the data. Plot these median ideal points by
  year using the variables `formerSovMedianIdealPt` and 
  `nonFormerSovMedianIdealPt` in the `unvotingYearAvg.csv` data set. Be
  sure to indicate 1989, the year of the fall of the Berlin
  Wall, on the graph.  Briefly comment on what you observe.

## Answer 5

```{r}
## Plot
plot(x = un.year.avg$Year, 
     y = un.year.avg$formerSovMedianIdealPt, 
     type = "l", ylim = c(-3, 3), xlab = "Year", ylab = "Ideal point")
lines(x = un.year.avg$Year,
      y = un.year.avg$nonFormerSovMedianIdealPt, 
      col = "red")
abline(v = 1989, lty = "dashed")
```

Before the fall of the Berlin Wall, the median Soviet state was far
below (less liberal than) the median non-Soviet UN country. However,
since the fall of the Berlin Wall, the median post-Soviet state has
actually been more liberal than the median non-post-Soviet UN country.
