---
title: 'Changing Minds on Gay Marriage: Revisited'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=FALSE,comment='')
```

In this exercise, we revisit the gay marriage study we analyzed previously. 
It is important to work on that exercise before answering the following questions.  

In May 2015, three
scholars reported several irregularities in the dataset used to
produce the results in the study.  This exercise is based on
  the unpublished report 'Irregularities in LaCour (2014)' by David
  Broockman, Joshua Kalla, and Peter Aronow. 

They found that the gay
marriage experimental data were statistically indistinguishable from
data in the Cooperative Campaign Analysis Project (CCAP), which
interviewed voters throughout the 2012 United States presidential
campaign. The scholars suggested that the CCAP survey data -- and not
the original data alleged to have been collected in the experiment --
were used to produce the results reported in the gay marriage
study. The release of a report on these irregularities ultimately led
to the retraction of the original article. In this exercise, we will
use several measurement strategies to reproduce the irregularities
observed in the gay marriage dataset.

To do so, we will use four CSV data files. The first three, 
`gayreshaped.csv`, `gayreshaped_study1.csv` 
and `gayreshaped_study2.csv`, are reshaped versions of the
original dataset in which every observation corresponds to a unique
respondent from both studies 1 and 2, study 1, and study 2, respectively. 
The variables in these files are:

-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `study`              Which study the data set is from (1 = Study1, 2 = Study2) 
 
 `treatment`          Five possible treatment assignment options 
 
 `therm1`             Survey thermometer rating of feeling towards gay couples in wave 1 (0--100) 
 
 `therm2`             Survey thermometer rating of feeling towards gay couples in wave 2 (0--100) 
 
 `therm3`             Survey thermometer rating of feeling towards gay couples in wave 3 (0--100) 
 
 `therm4`             Survey thermometer rating of feeling towards gay couples in wave 4 (0--100) 
-------------------------------------------------------------------------------

The last data file is the 2012 CCAP dataset alleged to have been used as the basis for
the gay marriage study results, `ccap2012.csv`.  The variables in the CCAP data are:

-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `caseid`             Unique respondent ID 
 
 `gaytherm`           Survey thermometer rating (0-100) of feeling towards gay couples 
-------------------------------------------------------------------------------

Note that a feeling thermometer measures how
warmly respondents feel toward gay couples on a 0-100 scale.

## Question 1

In the gay marriage study, researchers used seven waves of a
  survey to assess how lasting the persuasion effects were over time.
  One irregularity the scholars found is that responses across survey
  waves in the control group (where no canvassing occurred) had
  unusually high correlation over time.  What is the correlation
  between respondents' feeling thermometer ratings in waves 1 and 2
  for the control group in Study 1? To handle missing data, we should
  set the `use` argument of the `cor()` function to
  `"complete.obs"` so that the correlation is computed using
  only observations that have no missing data in any of these
  observations. To make our code simpler and easier to interpret, 
  we should also perform computations using a subset of study 1 that only 
  contains observations in the control group. In R, one way
  we can subset observations
  in a dataframe is by specifying a logical test with square brackets (`[]`)
  followed by a comma and a space (`, `), e.g., 
  `dataframe[variable.one == 1, ]`
  is a subset of `dataframe` only containing observations where
  $variable.one = 1$.  
  

## Answer 1

```{r}
gay1 <- read.csv("data/gayreshaped_study1.csv")
## Subset Study 1 Control Group
gay1.control <- gay1[gay1$treatment == "No Contact",]
## Compute correlation
cor(gay1.control$therm1, gay1.control$therm2, use = "complete.obs")
```

Given that correlation ranges from -1 to 1, the responses at the
two waves in Study 1 have a nearly perfect positive correlation.  This
result implies that the measured attitudes in this study were
unusually stable over time.  In typical panel studies in which the
same individual respondents are interviewed multiple times, we often
find lower correlations between responses over time because people are
prone to changing their responses in either systematic or
idiosyncratic ways across multiple waves of a survey.

## Question 2

Repeat the previous question, using Study 2 and comparing all
  waves within the control group.  Note that the `cor()` function
  can take a single dataframe with multiple variables as an argument
  and will compute correlations between all possible combinations
  of those variables. To handle missing data in this case, 
  we can set the `use` argument to `"pairwise.complete.obs"`. 
  This means that the
  `cor()` function uses all observations which have no missing
  values for a given pair of waves even if some of them have missing
  values in other waves. Briefly interpret the results. **Hint:**
  The function `data.frame()` takes individual variables as 
  arguments and joins them together into a single dataframe. We
  can use this function to create a dataframe consisting solely
  of "therm" variables and use it as an input for `cor()`.
  
## Answer 2

```{r}
gay2 <- read.csv("data/gayreshaped_study2.csv")
## Study 2 Control group
gay2.control <- gay2[gay2$treatment == "No Contact", ]
## Create dataframe with just thermometer ratings
gay2.therms <- data.frame(gay2.control$therm1, gay2.control$therm2, 
                          gay2.control$therm3, gay2.control$therm4)
## Compute correlation
cor(gay2.therms, use = "pairwise.complete.obs")
```

The results for Study 2 are consistent with what we found in Study 1.
All correlations are extremely high.

## Question 3

Most surveys find at least some *outliers* or individuals
  whose responses are substantially different from the rest of the
  data.  In addition, some respondents may change their responses
  erratically over time. Create a scatterplot to visualize the
  relationships between wave 1 and each of the subsequent waves in
  Study 2.  Use only the control group.  Interpret the results.

## Answer 3

```{r}
par(cex = 1.25)
plot(gay2.control$therm1, gay2.control$therm2,
     xlab = "Baseline", ylab = "Wave 2")
plot(gay2.control$therm1, gay2.control$therm3, 
     xlab = "Baseline", ylab = "Wave 3")
plot(gay2.control$therm1, gay2.control$therm4, 
     xlab = "Baseline", ylab = "Wave 4")
```

The figures show no respondents who exhibited changes in their feeling
thermometer responses that are out of step with the general pattern of
responses in the data. This is highly unusual for data that are
alleged to have been drawn from actual survey participants.

## Question 4

The researchers found that the data of the gay marriage study
  appeared unusually similar to the 2012 CCAP dataset even though they
  were supposed to be samples of completely different respondents. We
  use the data contained in `ccap2012.csv` to compare the samples. 
  To get a sense of the distribution of the samples, compute the 
  median and standard deviation for the 2012 CCAP feeling thermometer, 
  the wave 1 feeling thermometer from Study 1, and the wave 1 feeling 
  thermometer from Study 2. Briefly comment on the results.

## Answer 4

```{r}
## Load CCAP data
ccap <- read.csv("data/ccap2012.csv") 
## Compute medians and std. deviations
median(ccap$gaytherm, na.rm = TRUE)
median(gay1$therm1, na.rm = TRUE)
median(gay2$therm1, na.rm = TRUE)
sd(ccap$gaytherm, na.rm = TRUE)
sd(gay1$therm1, na.rm = TRUE)
sd(gay2$therm1, na.rm = TRUE)
```

The medians and standard deviations of all samples are (or are nearly)
the same. This suggests that the distributions of the CCAP as
well as gay marriage study, wave 1 data and gay marriage
study 1 and study 2, wave 1 data may be quite similar to one another. 
This is consistent with the claim that the study data were
not collected as described, and instead were generated
from the 2012 CCAP data.

## Question 5

Create a histogram of the 2012 CCAP feeling thermometer, the wave 1 feeling
  thermometer from Study 1, and the wave 1 feeling thermometer from
  Study 2 to help visualize their respective distributions.  Note that
  there are a large number of missing values in the CCAP
  data.  Consider how the missing data might have been recoded in the
  gay marriage study.  To facilitate comparison across histograms,
  use the `breaks` argument in the `hist` function to
  keep the bin sizes equal cross histograms.  Provide a brief substantive
  review of the results. **Hint:** The `is.na()` function might prove
  useful for this question. This function acts like a logical 
  test, returning `TRUE` for all values of its argument that are
  `NA` and returning `FALSE` for all values of its argument that are not `NA`.
  
## Answer 5

```{r}
par(cex = 1.25)

## Generate histograms
hist(gay1$therm1, freq = FALSE, breaks = 20, ylim = c(0, 0.05),
     xlab = "Feeling Thermometer", main = "Study 1, Baseline")
hist(gay2$therm1, freq = FALSE, breaks = 20, ylim = c(0, 0.05),
     xlab = "Feeling Thermometer", main = "Study 2, Baseline")
hist(ccap$gaytherm, freq = FALSE, breaks = 20, ylim = c(0, 0.05), 
     xlab = "Feeling Thermometer", main = "CCAP: without missing data")
## proportion of observations that are equal to 50
mean(gay1$therm1 == 50)
mean(gay2$therm1 == 50)
## Replace NAs with 50
ccap$gaytherm[is.na(ccap$gaytherm)] <- 50
hist(ccap$gaytherm, freq = FALSE, breaks = 20, ylim = c(0, 0.05), 
     xlab = "Feeling Thermometer", 
     main = "CCAP: with missing data as 50")
```

Comparing the two histograms from the gay marriage study to the
histogram of CCAP, we observe that the distributions are quite similar
except one bin in the gay marriage study histogram is not as tall as
the histograms of the gay marriage study.  A large fraction of
observations of the gay marriage data exactly equal 50.  Once we
recode missing values in the gay marriage study data as 50, the CCAP
histogram is indistinguishable from the gay marriage study histograms.


