---
title: "Sample - Summary - Homework"
author: "Jose Rico"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float : true
    theme: cosmo
    highlight: tango
---

```{r,  set.seed(1123581321), echo=F}
knitr::opts_chunk$set(cache=T, echo = T, warning=FALSE, message=FALSE, fig.height=4, fig.align='center') #Chunk default settings 
```

```{r, include=F, message=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(janitor)
library(infer)
library(kableExtra)
library(gridExtra)
library(scales)
library(tidyr)
library(mosaic)
```

## Assignment Overview

This will be your summative assignment for the sampling unit.  You will be;  

  - conducting a simulation
  - analyzing the simulations with statistics and visualizations
  - constructing confidence intervals
  
## Assignment Scenario

We will be analyzing a particular game, the rolling of three dice.  A winning rolls is when you get at least a match of two die faces.  For example; 1,1,4 would be a winning roll.   1,2,3 would be a losing roll.    

## Constructing a population data frame

Simulate rolling 100,000 times
```{r}
set.seed(1123581321)
roll1 = sample(1:6,100000, replace=T)
roll2 = sample(1:6,100000, replace=T)
roll3 = sample(1:6,100000, replace=T)
roll_sim_population <- data.frame(roll1, roll2, roll3) %>% 
  mutate(win = if_else(roll1==roll2 | roll1==roll3 | roll2==roll3, TRUE, FALSE))
```

## Assignment Tasks

### Sampling/Analysis/Narratives for different sample sizes
1. Take 1000 samples for each sample size 25, 100 and 400.

#### Taking 1000 samples, for n= 25, 100, 400
```{r}
set.seed(1123581321)
samples_n25_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 25, reps = 1000, replace = F)

samples_n100_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 100, reps = 1000, replace = F)

samples_n400_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 400, reps = 1000, replace = F)

```

2. Construct sampling distributions histograms for  sample size = 25.  

#### Calculate Summary Statistics for Each Sample (n=25)
```{r}
summary_n25_reps1000 <- samples_n25_reps1000 %>% 
  
          group_by(replicate) %>% 
  
          summarize(win = sum(win == "TRUE")) %>% 
  
          mutate(prop_win = win/25)  #### need to change divisor to match the sample size
```

#### Create histogram for distribution of 1000 samples for sample size (n=25).
```{r}
summary_n25_reps1000 %>% 
  
  ggplot(aes(prop_win)) +
  
  geom_histogram(binwidth = 0.01, 
                 color="white", 
                 fill="steelblue") +
  
  labs(x = "Proportion of Winning Rolls (Individual Sample)",
       title ="Distribution of Proportion of Winning Rolls for 1000 Samples \n(based on sample size of 25)") +
  
  
    scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,0.08))
```

2. Construct sampling distributions histograms for  sample size = 100.  

#### Calculate Summary Statistics for Each Sample (n=100)
```{r}
summary_n100_reps1000 <- samples_n100_reps1000 %>% 
  
          group_by(replicate) %>% 
  
          summarize(win = sum(win == "TRUE")) %>% 
  
          mutate(prop_win = win/100)  #### need to change divisor to match the sample size
```

#### Create histogram for distribution of 1000 samples for sample size (n=100).
```{r}
summary_n100_reps1000 %>% 
  
  ggplot(aes(prop_win)) +
  
  geom_histogram(binwidth = 0.01, 
                 color="white", 
                 fill="steelblue") +
  
  labs(x = "Proportion of Winning Rolls (Individual Sample)",
       title ="Distribution of Proportion of Winning Rolls for 1000 Samples \n(based on sample size of 100)") +
  
  
    scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,0.08))
```


3. Calculate (and Table) sampling distributions statistics for each sample size. (mean, sd, 2.5th and 97.5th percentiles)

```{r}
n25_dist_mean = mean(summary_n25_reps1000$prop_win)
n25_dist_sd = round(sd(summary_n25_reps1000$prop_win),3)
n25_bottom_2.5 = quantile(summary_n25_reps1000$prop_win,0.025)
n25_top_2.5 = quantile(summary_n25_reps1000$prop_win,0.975)

n100_dist_mean = mean(summary_n100_reps1000$prop_win)
n100_dist_sd = round(sd(summary_n100_reps1000$prop_win),3)
n100_bottom_2.5 = quantile(summary_n100_reps1000$prop_win,0.025)
n100_top_2.5 = quantile(summary_n100_reps1000$prop_win,0.975)
```


### Table of Statistics for Sampling Distributions

|  sample size  | mean   |  sd  |  bottom_2.5  |  top_2.5 |
|--------------|--------|--------|-----------|----------|
| 25  | `r n25_dist_mean` | `r n25_dist_sd` | `r n25_bottom_2.5`  | `r n25_top_2.5`  |
| 100  | `r n100_dist_mean` | `r n100_dist_sd` | `r n100_bottom_2.5`  | `r n100_top_2.5`  |




##### Prompt 1 - Explain what happens to the sampling distribution as the sample size increases from 25 to 100 to 400.

> **your answer goes here**

##### Prompt 2 - How would you answer the following question, What winning percentages would be unlikely for a sample of 25? 100? and 400?  If they are different, why?  If they are the same, why? 

> **your answer goes here**

4. Take a single sample for each sample size 25, 100 and 400 and construct a 95% confidence interval for proportion of wins for the population.

#### Take a sample of size 25 from our population

```{r}
set.seed(1123581321)
sample_n25 <- roll_sim_population %>% 
  sample_n(size = 25, replace = F )
```

#### Construct CI for single sample of n=25

```{r}
binom.test(~(win == "TRUE"), conf.level = 0.95, ci.method= "Wald", data = sample_n25) %>% 
  confint() %>% 
    kable() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

##### Prompt 3 - Compare your confidence intervals for different sample sizes.  How are they similar and/or different?

> **your answer goes here**

##### Prompt 4 - Interpret in context what your confidence interval for a sample size of 400 means?

> **your answer goes here**





