---
title: "Sample - Summary - Homework"
author: "Jose Rico"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float : true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F, message=FALSE}
library(readr)
library(tidyverse)
library(janitor)
library(infer)
library(kableExtra)
library(gridExtra)
library(scales)
library(tidyr)
```

## Assignment Overview

This will be your summative assignment for the sampling unit.  You will be;
  - conducting a simulation
  - analysizing the simulations with statistics and visualizations
  - constucting confidence intervals
  
## Assignment Scenario

We will be analyzing a particular game, the rolling of three dice.  A winning rolls is when you get at least a match of two die faces.  For example; 1,1,4 would be a winning roll.   1,2,3 would be a losing roll.    

## Constructing a population data frame

Simulate rolling 100,000 times
```{r}
roll1 = sample(1:6,100000, replace=T)
roll2 = sample(1:6,100000, replace=T)
roll3 = sample(1:6,100000, replace=T)
roll_sim_population <- data.frame(roll1, roll2, roll3) %>% 
  mutate(win = if_else(roll1==roll2 | roll1==roll3 | roll2==roll3, TRUE, FALSE))
```

## Assignment Tasks

### Sampling/Analysis/Narratives for different sample sizes
1. Take 1000 samples for each sample size 25, 100 and 400.
2. Construct sampling distributions histograms for each sample size.
3. Calculate (and Table) sampling distributions statistics for each sample size. (mean, sd, 2.5th and 97.5th percentiles)

##### Prompt 1 - Explain what happens to the sampling distribution as the sample size increases from 25 to 100 to 400.

> **your answer goes here**

##### Prompt 2 - How would you answer the following question, What winning percentages would be unlikely for a sample of 25? 100? and 400?  If they are different, why?  If they are the same, why? 

> **your answer goes here**

4. Take a single sample for each sample size 25, 100 and 400 and construct a 95% confidence interval for proportion of wins for the population.

##### Prompt 3 - Compare your confidence intervals for different sample sizes.  How are they similar and/or different?

> **your answer goes here**

##### Prompt 4 - Interpret in context what your confidence interval for a sample size of 400 means?

> **your answer goes here**





