---
title: "Sample - Summary - Homework"
author: "Jose Rico"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float : true
    theme: cosmo
    highlight: tango
---

```{r,  set.seed(1123581321), echo=F}
knitr::opts_chunk$set(cache=T, echo = T, warning=FALSE, message=FALSE, fig.height=4, fig.align='center') #Chunk default settings 
```

```{r, include=F, message=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(janitor)
library(infer)
library(kableExtra)
library(gridExtra)
library(scales)
library(tidyr)
library(mosaic)
```

## Assignment Overview

This will be your summative assignment for the sampling unit.  You will be;  

  - conducting a simulation (I have already done that for you)
  - analyzing the simulations with statistics and visualizations
  - constructing confidence intervals from sample statistics
  
## Assignment Scenario

We will be analyzing a particular game, the rolling of three dice.  A winning rolls is when you get at least a match of two die faces.  For example; 1,1,4 would be a winning roll.   1,2,3 would be a losing roll.    

## Constructing a population data frame

Simulate rolling 100,000 times
```{r}
set.seed(1123581321)
roll1 = sample(1:6,100000, replace=T)
roll2 = sample(1:6,100000, replace=T)
roll3 = sample(1:6,100000, replace=T)
roll_sim_population <- data.frame(roll1, roll2, roll3) %>% 
  mutate(win = if_else(roll1==roll2 | roll1==roll3 | roll2==roll3, TRUE, FALSE))
```

## Assignment Tasks

### Sampling/Analysis/Narratives for different sample sizes

1. Take 1000 samples for sample size 25
2. Construct sampling distributions histograms for sample size of 25.  
3. Calculate (and Table) sampling distributions statistics for each sample size. (mean, sd, 2.5th and 97.5th percentiles)

#### Sample Size of 25
```{r, echo=F}
set.seed(1123581321)
### First take 1000 samples with sample size of 25
samples_n25_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 25, reps = 1000, replace = F)

### Calculate Summary Statistics
summary_n25_reps1000 <- samples_n25_reps1000 %>% 
          group_by(replicate) %>% 
          summarize(win = sum(win == "TRUE")) %>% 
          mutate(prop_win = win/ 25)

### Create Histogram
summary_n25_reps1000 %>% 
  
  ggplot(aes(prop_win)) +
  
  geom_histogram(binwidth = 0.01, 
                 color="white", 
                 fill="steelblue") +
  
  labs(x = "Proportion of Winning Rolls (Individual Sample)",
       title ="Distribution of Proportion of Winning Rolls for 1000 Samples \n(based on sample size of 25)") +
  
    scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,0.08))

### calculation of values
s25_dist_mean = mean(summary_n25_reps1000$prop_win)
s25_dist_sd = round(sd(summary_n25_reps1000$prop_win),3)
s25_bottom_2.5 = quantile(summary_n25_reps1000$prop_win,0.025)
s25_top_2.5 = quantile(summary_n25_reps1000$prop_win,0.975)
```

#### Sample Size of 100
```{r, echo=F}
set.seed(1123581321)
### First take 1000 samples with sample size of 25
samples_n100_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 100, reps = 1000, replace = F)

### Calculate Summary Statistics
summary_n100_reps1000 <- samples_n100_reps1000 %>% 
          group_by(replicate) %>% 
          summarize(win = sum(win == "TRUE")) %>% 
          mutate(prop_win = win/ 100)

### Create Histogram
summary_n100_reps1000 %>% 
  
  ggplot(aes(prop_win)) +
  
  geom_histogram(binwidth = 0.01, 
                 color="white", 
                 fill="steelblue") +
  
  labs(x = "Proportion of Winning Rolls (Individual Sample)",
       title ="Distribution of Proportion of Winning Rolls for 1000 Samples \n(based on sample size of 100)") +
  
    scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,0.08))

### calculation of values
s100_dist_mean = mean(summary_n100_reps1000$prop_win)
s100_dist_sd = round(sd(summary_n100_reps1000$prop_win),3)
s100_bottom_2.5 = quantile(summary_n100_reps1000$prop_win,0.025)
s100_top_2.5 = quantile(summary_n100_reps1000$prop_win,0.975)
```

#### Sample Size of 400
```{r, echo=F}
set.seed(1123581321)
### First take 1000 samples with sample size of 25
samples_n400_reps1000 <- roll_sim_population %>%
        rep_sample_n(size = 400, reps = 10000, replace = F)

### Calculate Summary Statistics
summary_n400_reps1000 <- samples_n400_reps1000 %>% 
          group_by(replicate) %>% 
          summarize(win = sum(win == "TRUE")) %>% 
          mutate(prop_win = win/400)

### Create Histogram
summary_n400_reps1000 %>% 
  
  ggplot(aes(prop_win)) +
  
  geom_histogram(binwidth = 0.005, 
                 color="white", 
                 fill="steelblue") +
  
  labs(x = "Proportion of Winning Rolls (Individual Sample)",
       title ="Distribution of Proportion of Winning Rolls for 1000 Samples \n(based on sample size of 400)") +
  
    scale_x_continuous(limits = c(0, 1), breaks = seq(0,1,0.08))

### calculation of values
s400_dist_mean = mean(summary_n400_reps1000$prop_win)
s400_dist_sd = round(sd(summary_n400_reps1000$prop_win),3)
s400_bottom_2.5 = quantile(summary_n400_reps1000$prop_win,0.025)
s400_top_2.5 = quantile(summary_n400_reps1000$prop_win,0.975)
```

|  sample size  | mean   |  sd  |  bottom_2.5  |  top_2.5 |
|--------------|--------|--------|-----------|----------|
| 25  | `r s25_dist_mean` | `r s25_dist_sd` | `r s25_bottom_2.5`  | `r s25_top_2.5`  |
| 100  | `r s100_dist_mean` | `r s100_dist_sd` | `r s100_bottom_2.5`  | `r s100_top_2.5`  |
| 400 | `r s400_dist_mean` | `r s400_dist_sd` | `r s400_bottom_2.5`  | `r s400_top_2.5`  |


u

> **your answer goes here**

4. Take a single sample for each sample size 25, 100 and 400 and construct a 95% confidence interval for proportion of wins for the population.

n=25
```{r, echo=F}
#set.seed(1123581321)

n = 25

sample1 <- roll_sim_population %>% 
  sample_n(size = n, replace = F ) 

stat_sample1 <- sample1 %>% 
  
      summarize(win = sum(win == "TRUE"),
                prop_win = win/ n,
                se = sqrt(prop_win*(1-prop_win)/n),
                me = qnorm(0.975) * se,
                lower = prop_win - me,
                upper = prop_win + me)

kable(stat_sample1) %>% 
    kable_styling(bootstrap_options = "striped", 
                  full_width = F, position = "left")
### this technique does not use the mosaic package - that's why Ish - it worked.
```

n= 100
```{r, echo=F}
#set.seed(1123581321)

n = 100

sample1 <- roll_sim_population %>% 
  sample_n(size = n, replace = F ) 

stat_sample1 <- sample1 %>% 
  
      summarize(win = sum(win == "TRUE"),
                prop_win = win/ n,
                se = sqrt(prop_win*(1-prop_win)/n),
                me = qnorm(0.975) * se,
                lower = prop_win - me,
                upper = prop_win + me)

kable(stat_sample1) %>% 
    kable_styling(bootstrap_options = "striped", 
                  full_width = F, position = "left")
```

n= 400
```{r, echo=F}
#set.seed(1123581321)
n = 400

sample1 <- roll_sim_population %>% 
  sample_n(size = n, replace = F ) 

stat_sample1 <- sample1 %>% 
  
      summarize(win = sum(win == "TRUE"),
                prop_win = win/ n,
                se = sqrt(prop_win*(1-prop_win)/n),
                me = qnorm(0.975) * se,
                lower = prop_win - me,
                upper = prop_win + me)

kable(stat_sample1) %>% 
    kable_styling(bootstrap_options = "striped", 
                  full_width = F, position = "left")
```

##### Prompt 3 - Compare your confidence intervals for different sample sizes.  How are they similar and/or different?

> **your answer goes here**

##### Prompt 4 - Interpret in context what your confidence interval for a sample size of 400 means?

> **your answer goes here**


```{r}
ab <- 12
cd <- 24
ef <- 36
```

| col1 | col2 | col3 |
|-----|----|----|
| ab | `r ab` |  |
| cd | `r cd` | i did it! |


## In Class - Generation of Population to Sample from...
```{r}
pop <- data.frame(individual = sample(c("satisfied", "not satisfied"), size = 1000000, replace=T, prob=c(0.37, 0.63)))
```






