---
title: "Bias in Self-reported Turnout"
output:
  html_document: default
  pdf_document: default
---

Surveys are frequently used to measure political behavior such as
voter turnout, but some researchers are concerned about the accuracy
of self-reports.  In particular, they worry about possible *social
desirability bias* where in post-election surveys, respondents who did
not vote in an election lie about not having voted because they may
feel that they should have voted.  Is such a bias present in the
American National Election Studies (ANES)?  The ANES is a nation-wide
survey that has been conducted for every election since 1948.  The
ANES conducts face-to-face interviews with a nationally representative
sample of adults.  The table below displays the names and descriptions
of variables in the `turnout.csv` data file.

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `year`               Election year
 
 `VEP`                Voting Eligible Population (in thousands) 
 
 `VAP`                Voting Age Population (in thousands) 
 
 `total`              Total ballots cast for highest office (in thousands) 
 
 `felons`             Total ineligible felons (in thousands) 
 
 `noncitizens`        Total non-citizens (in thousands) 
 
 `overseas`           Total eligible overseas voters (in thousands)
 
 `osvoters`           Total ballots counted by overseas voters (in thousands)
--------------------------------------------------------------------------------

We will also make use of derivative data files `pres_turnout.csv`, `mid_turnout.csv`, 
`earlier_turnout.csv`, and `later_turnout.csv`.

## Question 1

Load the data into R and check the dimensions of the data.
  How many observations are
  there?  Using the function `head()` and the argument `n`, list
  the range of years covered by the data. What type are all the
  variables in the dataframe?

## Answer 1

```{r}
turnout <- read.csv("data/turnout.csv")
dim(turnout)
head(turnout, n = 14)
```

There are `r dim(turnout)[1]` observations in this data set.  The
data covers the time period from
`r min(turnout$year)` to `r max(turnout$year)`. All variables in
the dataframe have integer values.

## Question 2

Calculate the turnout rate based on the voting age population or
  VAP and add it to the data as a new variable. Note that for this data set, 
  we must add the total number of
  eligible overseas voters since the *VAP* variable does not
  include these individuals in the count. Next, calculate the turnout
  rate using the voting eligible population or VEP and add it to the
  data as a new variable.  What difference
  do you observe? Make sure to leave comments identifying the
  calculations. **Hint:** We can use the function `View()` to
  browse the newly created variables.
  
## Answer 2

```{r}
## calculate turnout rate based on VAP
turnout$VAPtr <- turnout$total/(turnout$VAP + turnout$overseas) * 100
turnout$VAPtr
## calculate turnout rate based on VEP
turnout$VEPtr <- turnout$total/turnout$VEP * 100
turnout$VEPtr
## View(turnout)
```

We find that the VAP turnout rate is consistently lower than the VEP 
turnout rate. This result is due to the fact that the VAP includes 
numerous individuals who are ineligible to vote.

## Question 3

Compute the difference between VAP and ANES estimates of turnout
  rate.  How big is the difference on average?  Conduct the same 
  comparison for the VEP and ANES estimates of voter turnout. Make sure to 
  leave comments identifying the calculations.
  Briefly comment on the results.

## Answer 3

```{r}
## compare VAP and ANES
turnout$diffVAP <- turnout$ANES - turnout$VAPtr
mean(turnout$diffVAP)
## compare VEP and ANES
turnout$diffVEP <- turnout$ANES - turnout$VEPtr
mean(turnout$diffVEP)
```

On average, the ANES turnout rate is `r round(mean(turnout$diffVAP), 2)`
percentage points higher than the VAP turnout rate. Comparatively, we
find that on average, the ANES turnout rate is
`r round(mean(turnout$diffVEP), 2)` percentage points higher than the VEP
turnout rate. The ANES
turnout rate is higher than both the VAP and VEP turnout
rates. However, because the VAP turnout rate is consistently lower
than the VEP turnout rate, we find that the difference between the VEP
and ANES turnout rates is less than the difference between the VAP and
ANES turnout rates.

## Question 4

Presidential elections and midterm elections occur every four years,
  staggered by two years with respect to each other. Let's investigate 
  differences in midterm and
  presidential elections by loading the data 
  from `pres_turnout.csv` and `mid_turnout.csv`.
  Compute and compare the mean VEP turnout rate with the ANES turnout rate
  separately for presidential elections and midterm elections.  Note
  that the data set excludes the year 2006. Does the bias of the ANES
  vary across election types? 

## Answer 4

```{r}
## presidential
pres_turnout <- read.csv("data/pres_turnout.csv")
pres_turnout$VEPtr <- pres_turnout$total/pres_turnout$VEP * 100
mean(pres_turnout$ANES - pres_turnout$VEPtr)
## midterm
mid_turnout <- read.csv("data/mid_turnout.csv")
mid_turnout$VEPtr <- mid_turnout$total/mid_turnout$VEP * 100
mean(mid_turnout$ANES - mid_turnout$VEPtr)
```

On average, the bias is higher among presidential elections than
midterm elections. The ANES turnout rate is on average
`r round(mean(pres_turnout$ANES - pres_turnout$VEPtr), 2)` 
percentage points higher than the VEP
turnout rate in presidential elections.  In midterm elections, the
ANES turnout rate is on average `r round(mean(mid_turnout$ANES - mid_turnout$VEPtr), 2)`
percentage points higher than the VEP turnout rate.

## Question 5

Load data from `earlier_turnout.csv` and `later_turnout.csv` 
  so as to separately examine the first and second halves 
  of the year range of the original `turnout.csv` dataset.
  Calculate the mean difference between the VEP
  turnout rate and the ANES turnout rate
  within each period.  Has the bias of the ANES increased over time?

## Answer 5

```{r}
earlier_turnout <- read.csv("data/earlier_turnout.csv") # first half
later_turnout <- read.csv("data/later_turnout.csv") # second half
## comparison
earlier_turnout$VEPtr <- earlier_turnout$total/earlier_turnout$VEP * 100
mean(earlier_turnout$ANES - earlier_turnout$VEPtr)
later_turnout$VEPtr <- later_turnout$total/later_turnout$VEP * 100
mean(later_turnout$ANES - later_turnout$VEPtr)
```

We find that the bias of the ANES has increased over
time. For the first half of election years, the ANES turnout rate is
on average `r round(mean(earlier_turnout$ANES - earlier_turnout$VEPtr), 2)` percentage points
higher than the VEP turnout rate. Comparatively, the ANES is on
average `r round(mean(later_turnout$ANES - later_turnout$VEPtr), 2)` percentage points higher
than the VEP turnout rate among the later election years.

## Question 6

The ANES does not interview overseas voters and
  prisoners. Calculate an adjustment to the VAP turnout
  rate. Begin by subtracting the total number of ineligible felons and
  non-citizens from the VAP to calculate an adjusted VAP. Next,
  calculate an adjusted VAP turnout rate, taking care to subtract the
  number of overseas ballots counted from the total ballots.
  Compare the mean adjusted VAP, mean unadjusted VAP, and mean VEP to
  the ANES turnout rate. Briefly discuss the results.
  
## Answer 6

```{r}
turnout$adj.VAP <- turnout$VAP - (turnout$felons + turnout$noncit)
turnout$adj.VAPtr <- (turnout$total - turnout$osvoters) / turnout$adj.VAP * 100
turnout$adj.diff <- turnout$ANES - turnout$adj.VAPtr
## compare adjusted difference to non-adjusted VAP as well as VEP
mean(turnout$adj.diff)
mean(turnout$diffVAP)
mean(turnout$diffVEP)
```

The mean adjusted VAP turnout rate is closer to the ANES turnout rate than
the non-adjusted VAP or VEP turnout rates. The mean adjusted VAP
turnout rate is `r round(mean(turnout$adj.diff), 2)` percentage points lower
than the ANES turnout rate. The non-adjusted VAP turnout rate is
`r round(mean(turnout$diffVAP), 2)` percentage points lower than the ANES
turnout rate while the VEP turnout rate is `r round(mean(turnout$diffVEP), 2)` 
percentage points lower.  In general it appears that the ANES
overestimates turnout rates. However, by adjusting the VAP rate to
reflect the conditions of the ANES survey, we reduce the difference in
estimated turnout rates.
