---
title: "Political Efficacy in China and Mexico"
output:
  html_document: default
  pdf_document: default
---

In 2002, the World Health Organization conducted a survey of two
provinces in China and three provinces in Mexico. This
exercise is based on: 

> Gary King, Christopher J. L. Murray, Joshua A. Salomon, and Ajay Tandon. (2004). '[Enhancing the Validity and Cross-Cultural Comparability of Measurement in Survey Research.]( https://doi.org/10.1017/S000305540400108X)' *American Political Science Review*, 98:1 (February), pp.191-207.
  
In this exercise we analyze respondents' views on their own political efficacy.  First, the following self-assessment question was asked.

>  How much say do you have in getting the government to address issues that interest you?
>
>  (5) Unlimited say, (4) A lot of say, (3) Some say, (2) Little say, (1) No say at all.

After the self-assessment question, three vignette questions were
asked.

>  [Alison] lacks clean drinking water. She and her neighbors are supporting an opposition candidate in the forthcoming elections that has promised to address the issue. It appears that so many people in her area feel the same way that the opposition candidate will defeat the incumbent representative.
>
>  [Jane] lacks clean drinking water because the government is pursuing an industrial development plan. In the campaign for an upcoming election, an opposition party has promised to address the issue, but she feels it would be futile to vote for the opposition since the government is certain to win.
>
>  [Moses] lacks clean drinking water. He would like to change this, but he can't vote, and feels that no one in the government cares about this issue. So he suffers in silence, hoping something will be done in the future.

The respondent was asked to assess each vignette in the same manner as
the self-assessment question.

>  How much say does ['name'] in getting the government to address issues that interest [him/her]?
>
>  (5) Unlimited say, (4) A lot of say, (3) Some say, (2) Little say, (1) No say at all.

where ['name'] was replaced with either Alison, Jane, or Moses.

The data set we analyze `vignettes.csv` contains the following variables:

-------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `self`               Self-assessment response 
 
 `alison`             Response on Alison vignette 
 
 `jane`               Response on Jane vignette 
 
 `moses`              Response on Moses vignette
 
 `china`              1 for China and 0 for Mexico  
 
 `age`                Age of respondent in years 
------------------------------------------------------------------------------

In the analysis that follows, we assume
that these survey responses can be treated as numerical values.  For
example, `Unlimited say` = 5, and `Little say` = 2.  

This approach is not appropriate if, for example, the
difference between `Unlimited say` and `A lot of say` is 
not the same as the difference between `Little say` and `No say at all`.  However, relaxing this
assumption is beyond the scope of this chapter.

## Question 1

We begin by analyzing the self-assessment question.  Create tables
  of proportions of responses separately for China and Mexico.
  Briefly comment on the results.
  In addition, compute the mean 
  response for each country.  Which country's
  respondents seemt to have a higher degree of
  political efficacy?  Does this seem plausible when Mexican citizens 
  voted out out the Institutional Revolutionary Party (PRI) 
  after more than 70 years or rule and Chinese citizens have not been able to vote
  in a fair election to date? **Hint:** It might be helpful
  to create separate dataframes for respondents from China and Mexico.
  In R, one way we can subset observations
  in a dataframe is by using the function `subset()`.
  `subset()` takes two arguments, a dataframe followed by 
  logical test, and returns
  a subsetted dataframe where all observations pass the given
  logical test.

## Answer 1

```{r}
## Load data and create su
vig <- read.csv("data/vignettes.csv")
vig_chi <- subset(vig, china == 1)
vig_mex <- subset(vig, china == 0)
## China proportion table
prop.table(table(vig_chi$self))
## Mexico proportion table
prop.table(table(vig_mex$self))
## Average responses
mean(vig_chi$self) 
mean(vig_mex$self) 
```

Respondents in China appear to have a higher level of efficacy than
respondents in Mexico but this does not conform to what we know about the
ability of citizens to influence government. We would have expected Mexican citizens to show higher levels of
political efficacy. For example, citizens in
Mexico are able to vote out the ruling party, while
Chinese citizens are not. This could be a result of misreporting,
although we will need to do more to investigate this possibility.

## Question 2

We examine the possibility that any difference in the levels of
  efficacy between Mexican and Chinese respondents is due to the
  difference in their age distributions.  Create histograms for the
  age variable separately for Mexican and Chinese respondents.  Add a
  vertical line representing the median age of the respondents for
  each country.  What differences in age distribution do
  you observe between the two countries?  Answer this by
  interpreting each plot.

## Answer 2

```{r}
## Plots for China
hist(vig_chi$age, freq = FALSE,
     xlab = "Years of Age", main = "Distribution of Respondent's Age - China")
chi_age_median <- median(vig_chi$age)
abline(v = chi_age_median, 
       lty = "dashed", lwd = 2, col = "blue")
text("median", x = chi_age_median, y = 0.03, pos = 4, col = "blue")
## Plots for Mexico
hist(vig_mex$age, freq = FALSE,     
     xlab = "Years of Age", main = "Distribution of Respondent's Age - Mexico")
mex_age_median <- median(vig_mex$age)
abline(v = mex_age_median,
      lty = "dashed", lwd = 2, col = "blue")
text("median", x = mex_age_median, y = 0.03, pos = 4, col = "blue")
```

The histograms show that the Chinese respondents were older than their
Mexican counterparts. Thus, the differences we found in the previous questions 
may possibly result from differences in the age of Chinese and Mexican respondents.

## Question 3

One problem of the self-assessment question is that survey
  respondents may interpret the question differently.  For example,
  two respondents who choose the same response category may be facing quite
  different political situations and hence may interpret `A lot of say` 
  differently.  To address this problem, we rank a
  respondent's answer to the self-assessment question relative to the
  same respondent's answer to a vignette question.  Compute the
  proportion of respondents, again separately for China and Mexico,
  who ranks themselves (according to the self-assessment question) as
  having less say in the government's decisions than Moses (the last
  vignette).  To help perform this computation, employ the relational operator `<`, 
  which can be used to test whether or not the observations of a variable 
  are less than a particular value. How does the result of this analysis
  differ from that of the previous analysis?  Give a brief 
  interpretation of the result.
  
## Answer 3

```{r}
## China
mean(ifelse(vig_chi$self - vig_chi$moses < 0, 1, 0))
## Mexico
mean(ifelse(vig_mex$self - vig_mex$moses < 0, 1, 0))
```

This result differs from the previous
result. 
`r round(mean((vig_chi$self - vig_chi$moses) < 0) * 100)`
percent of Chinese citizens rate themselves below Moses in political
efficacy, while only
`r round(mean((vig_mex$self - vig_mex$moses) < 0) * 100)` percent
of Mexican respondents do the same.  This suggests that while Chinese
citizens may subjectively feel they have `A lot of say`
based on their interpretation of what this category means, they also
recognize their actual level of say is low compared to Moses in the
vignette.

## Question 4
  
Restrict the data to survey respondents who ranked these three vignettes
in the expected order (i.e., `Alison` $\ge$ `Jane` $\ge$
`Moses`).  Now create a variable that represents how respondents
rank themselves relative to each vignette.  This variable should
be equal to 1 if a respondent ranks themself lower than
`Moses`, 2 if ranked the same as `Moses` or higher than
`Moses` but lower than `Jane`, 3 if ranked the same as `Jane`
or higher than `Jane` but lower than `Alison`, and 4 if ranked as high as `Alison` or higher. 
Note that the relational operator `>=` can be used to test whether or not
the observations of a variable are greater than or equal to a particular value.
Create tables of proportions for this new variable for China and Mexico, respectively.
Also, compute the mean value of this new variable separately for China and
Mexico.  Give a brief interpretation of the result by comparing
these results with those obtained in Question 1.

## Answer 4

```{r}
## Subset of respondents who correctly ordered the vignettes
cvig <- subset(subset(vig, alison >= jane), jane >= moses)
## A new variable using the vignette responses
cvig$ranking <- ifelse(cvig$self >= cvig$alison, 4,
                       ifelse(cvig$self >= cvig$jane, 3,
                              ifelse(cvig$self >= cvig$moses, 2, 1)))
## Re-split into countries
cvig_chi <- subset(cvig, china == 1)
cvig_mex <- subset(cvig, china == 0)
## China
prop.table(table(cvig_chi$ranking))
mean(cvig_chi$ranking)
## Mexico
prop.table(table(cvig_mex$ranking))
mean(cvig_mex$ranking)
```

Mexican citizens have higher levels of efficacy based on this new
measure. This differs from the first result, which suggests that
Chinese citizens had greater efficacy. Under this alternative measure,
most Chinese citizens rank themselves lower than Moses, while more
than three-quarters of Mexican citizens give themselves higher
ratings. This demonstrates the importance of taking into account how
different groups of respondents may understand rating scales in
different ways.  By anchoring respondents based on how they compare to
the vignette, we can recover a more meaningful comparison of efficacy
between the citizens in China and Mexico.

## Question 5

Is the problem identified above more or less severe among older
respondents when compared to younger ones?  Consider the previous
question for those who are 40 years or older and those
who are younger than 40 years.  Do your conclusions between these 
two groups of respondents?
Relate your discussion to your finding for Question 2.

## Answer 5

```{r}
## Create subsets
cvig_mex_above40 <- subset(cvig_mex, age >= 40)
cvig_mex_below40 <- subset(cvig_mex, age < 40)
cvig_chi_above40 <- subset(cvig_chi, age >= 40)
cvig_chi_below40 <- subset(cvig_chi, age < 40)
## China self assessment, above and below 40
prop.table(table(cvig_chi_above40$self))
prop.table(table(cvig_chi_below40$self))
mean(cvig_chi_above40$self) - mean(cvig_chi_below40$self)
## China ranking, above and below 40
prop.table(table(cvig_chi_above40$ranking))
prop.table(table(cvig_chi_below40$ranking))
mean(cvig_chi_above40$ranking) - mean(cvig_chi_below40$ranking)
## Mexico self assessment, above and below 40
prop.table(table(cvig_mex_above40$self))
prop.table(table(cvig_mex_below40$self))
mean(cvig_mex_above40$self) - mean(cvig_mex_below40$self)
## Mexico ranking, above and below 40
prop.table(table(cvig_mex_above40$ranking))
prop.table(table(cvig_mex_below40$ranking))
mean(cvig_mex_above40$ranking) - mean(cvig_mex_below40$ranking)
```

Both age groups appear to exhibit the same pattern, so the difference in age
distribution found earlier cannot explain this problem.
This suggests that the bias in self-assessment applies to both young and old survey respondents.
