---
title: "HW 4 Explore a Single Variable - Guide"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    css: ../css/sandstone.css
  pdf_document: default
---


```{r set chunk option, echo=F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE, fig.height=4, fig.align='center') #Chunk default settings 
```

```{r library packages}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
```

```{r set ggplot2 parameters}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load data}
load("/cloud/project/data/demo_addhealth_clean.Rdata")

```


# Table of Variables

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
RACE  |  race_c  | subject's race
BIO_SEX4 | gender | subject's gender
--  |  age | subject's age
H4ED2 | college_deg | Dd you complete college (bachelor's degree)?
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4MH23 | too_tired | (During the past seven days:) You felt that you were too tired to do things.
H4LM19 | hrs_worked_week | How many hours a week (do/did) you usually work at this job?

# Examing categorical variable 
## 1st Categorical `vigorous_activity`

**Numerical Summary Table**  

```{r summary table cat 1}
clean %>%
  tabyl(vigorous_activity) %>%
  adorn_pct_formatting(digits=1) %>%
  adorn_totals()%>%
  kable() %>%
  kable_styling(full_width = F, position="center")
```
, y= ..prop.., group=1
fill="steelblue", width=0.5)+
  ggtitle("In the past 24 hours, have you participated in vigorous physical activity
  long enough to work up a sweat, get your heart thumping, or get out of breath?") +
  
**Bar Plot - Frequency of Levels**

```{r bar plot cat 1}
clean %>% 
  #filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  ggplot(aes(x=vigorous_activity, y= ..prop.., group=1)) +
    geom_bar(fill="steelblue", width=0.5) +
      ggtitle("In the past 24 hours, have you participated in vigorous physical activity long 
enough to work up a sweat, get your heart thumping, or get out of breath?") +
      xlab("subject's response")
  
```

**Narrative summary**  
<div id = "narrative">
The participants in the **addhealth** study were asked, "*In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?*" **5111** out of **5114** provided a "yes" or "no" response, with approximately **40%** of the participant responding **yes**.  Approximately **2 out of 5** participants have participated in a vigorous activity in the past 24 hours.
</div>

# Examining numerical variable 
## 1st Numerical `hrs_work_week`

**Numerical Summary Table** 

```{r summarize num 1}
hours <- clean %>% 
  filter(hrs_work_week < 121) %>% 
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = mean(hrs_work_week),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = sd(hrs_work_week))

kable(round(hours,1)) %>% 
kable_styling(full_width = F, position="center")
```


**Histogram - visualization of the distribution**

```{r hist num 1}
clean %>%
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(x=hrs_work_week)) + 
    geom_histogram(aes(y=..density..), color="white", fill="steelblue", binwidth = 5) + 
      xlab("hours per week") +
      ggtitle("How many hours a week (do/did) you usually work at this job?") 
```

**Boxplot - visualization of 5 number summary**

```{r boxplot num 1}
clean %>%
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(y=hrs_work_week, x=0)) + 
    geom_boxplot(color="steelblue", width=0.6) + 
      ggtitle("How many hours a week (do/did) you usually work at this job?")+
  coord_flip()
```



**Narrative Summary**
<div id = "narrative">

Participants were asked, in reference to their curren or most recent job (which they work more than 10 hours/week), "*How many hours a week (do/did) you usually work at this job?*"  

* **5023** out of **5114** responded with hours between 10 and 120.  
  
* **82** participants were a **legitimate skip** since they have not worked a paying job for at least 10 hours a week, they are not represented here.  

The center of the distribution is best described with the **mode** and **median** of 40 hours/week.   There is considerable overall variation as hours vary from 10 to 120, although the middle 50% of particpants are between 40 and 45 hours, for an interquartile range (**IQR**) of 5 hours.  Weekly work hours is *slightly skewed to the **right*** (since outliers extend, to the right, to 120 hours).  This skew also results in the **mean** being greater than the **median**, which is a typically characteristic of right skewed distributions. 
</div>





