---
title: "explore-variables"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    css: ../css/sandstone.css
  pdf_document: default
---


```{r set chunk option, echo=F}
knitr::opts_chunk$set(echo = F, warning=FALSE, message=FALSE, fig.height=4, fig.align='center') #Chunk default settings 
```

```{r lbrary packages}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
```


```{r set ggplot2 parameters}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load data}
load("/cloud/project/data/demo_addhealth_clean.Rdata")
```


## Table of Variables

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
RACE  |  race_c  | subject's race
BIO_SEX4 | gender | subject's gender
--  |  age | subject's age
H4ED2 | college_deg | Dd you complete college (bachelor's degree)?
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4MH23 | too_tired | (During the past seven days:) You felt that you were too tired to do things.
H4LM19 | hrs_worked_week | How many hours a week (do/did) you usually work at this job?

## Examing categorical variable 
### 1st Categorical `vigorous_activity`

**Numerical Summary Table**  

```{r}
summary(clean$vigorous_activity)
```


```{r summary table cat 1}
dt <- clean %>%
  #filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  tabyl(vigorous_activity) %>%
  adorn_pct_formatting(digits=1) %>%
  adorn_totals()
dt %>%
  kable() %>%
  kable_styling(full_width = F, position="center")
```

**Bar Plot - Frequency of Levels**

```{r bar plot cat 1}
clean %>% 
  #filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  ggplot(aes(x=vigorous_activity, y= ..prop.., group=1))+
  geom_bar(fill="steelblue", width=0.5)+
  ggtitle("In the past 24 hours, have you participated in vigorous physical activity
  long enough to work up a sweat, get your heart thumping, or get out of breath?") +
  xlab("subject's response")
```

## Examining numerical variable 
### 1st Numerical `hrs_work_week`

**Numerical Summary Table** 

```{r summarize num 1}
hours <- clean %>% 
  filter(hrs_work_week < 121) %>% 
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = mean(hrs_work_week),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = sd(hrs_work_week))

kable(round(hours,2)) %>% 
kable_styling(full_width = F, position="center")
```


**Histogram - visualization of the distribution**

```{r hist num 1}
clean %>%
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(x=hrs_work_week)) + 
    geom_histogram(aes(y=..density..), color="white", fill="steelblue", binwidth = 5) + 
      xlab("hours per week") +
      ggtitle("How many hours a week (do/did) you usually work at this job?") 
```

**Boxplot - visualization of 5 number summary**

```{r boxplot num 1}
clean %>%
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(y=hrs_work_week, x=0)) + 
    geom_boxplot(color="steelblue", width=0.6) + 
      ggtitle("How many hours a week (do/did) you usually work at this job?") +
  coord_flip()
```












