---
title: "Describing distributions of data"
output:
  pdf_document: default
  word_document:
    toc: yes
  html_document:
    highlight: tango
    theme: readable
    toc: yes
    toc_float: yes
css: ../css/sandstone.css
---

# Assignment Overview

There are a variety of conventional ways to visualize data - tables, histograms, bar graphs, etc. The purpose is always to examine the distribution of variables related to your research question. You will create a plot, follow up each graphic with a table of summary statistics (for quantitative variables) or frequency and proportion table (for categorical), and then a summary paragraph that brings it all together. 

# Instructions

* Use the template provided: [[RMD]](hw04_univ_graphing_template.Rmd). 
* Completely describe 2 categorical and 2 quantitative variables using 
    - A table of summary statistics,
    - An appropriate plot with titles and axes labels, 
    - A short paragraph description in full complete English sentences. 

## Guidiance

* What is the trend in the data? What exactly does the chart show? (Use the chart title to help you answer this question)
* Describe the shape:
    - Symmetry/Skewness - Is it symmetric, skewed right, or skewed left?
    - Modality - Is it uniform, unimodal, or bimodal?             
* Describe the spread:
    - Variability - What is the approximate range of the data (x-axis)?
    - Does the variable have a lot of variability in the data (visually, are the participants responded to many different responses or mainly just one)?
* Describe the center: What is the mean/median/midpoint of the data? (Pick one or two). Don't 
* Describe the outliers (note: there may not be any for every graph):
    - Are there any outliers for the variable?
    - If yes, are these true outliers or false (due to data management or input error) outliers?
* Reread your explanation for context grammer, spelling and common sense. 

# Submission

1. Upload the final PDF to `04 Univariate Graphing` folder in Google Drive with the file name: `univ_graphing_userid.pdf` by the due date. 

\newpage 

# Example
This example uses the `mpg` data set from the `ggplot2` package. 
```{r, warning=FALSE, message=FALSE}
library(sjPlot) # For plotting using the sjp.frq() function
library(ggplot2) # For plotting using ggplot() function
library(knitr) # To make nice tables
library(descr) # For plotting using the freq() function
mpg <- ggplot2::mpg # you would load() your clean data here
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE) # options to suppress warnings and messages in subsequent code chunks
```

## Example of a basic-level answer for a categorical variable
This example shows a draft style plot, direct computer output showing/copied. Poor grammar and/or sentence structure, no attempt at explaining what the variable means, extra unnecessary or incorrect information included. Typos. 

> class

```{r}
freq(mpg$class)
```

> theres more suvs than compacts. 2% are 2seaters. there are 5 2seaters 47 cmpact 41 midize 11 minivans 33 pickups 35% subcompacts, 62 suv and 234 total cars. 

## Example of a proficient-level answer for a categorical variable
This example has a cleaned up plot, full English sentences, useful text formatting of variable names and levels. Explained what the variable was named and what it measured. 

> The `class` variable from the `mpg` data set is a catgorical variable that describes the type of vehicle being measured. Some levels of this categorical variable include _compact_, _pickup_ and _suv_. 

```{r, fig.height=4, fig.align='center'}
set_theme(base = theme_classic())
sjp.frq(mpg$class)
```

> Sub compact cars are the most frequently reported type of car, making up over one-quarter (26.5%) of the cars in this data set with n=62 cars represented. The least represented car is a compact car with n=5 (2.1%) records. 

\newpage

## Example of a basic-level answer for a quantiative variable
No english description provided, no verbal explanation of what information was gained from these plots. 
```{r}
ggplot(mpg, aes(cty)) + geom_histogram()
summary(mpg$cty)
```

## Example of a proficient-level answer for a quantitative variable
This example uses a histogram with overlaid density curve, and a boxplot to understand the shape, location and to look for outliers. Table of summary statistics present in a nicely formatted way, digits rounded appropriately. Plot cleaned up with appropriate axis and titles.  

> The `cty` variable records the miles per gallon (mpg) achieved during city driving. This is a quantititative numeric variable. 

```{r, fig.height=3, fig.width=4, fig.align='center'}
ggplot(mpg, aes(x=cty)) + geom_histogram(aes(y=..density..), 
                                         fill="grey", binwidth = 2) + 
      geom_density() + xlab("MPG") +
      ggtitle("City miles per gallon (MPG)") 
boxplot(mpg$cty)
kable(t(c(summary(mpg$cty), sd=sd(mpg$cty))), digits=1)
```

> The MPG in the city ranges from 9 to 35, unimodal and is slightly skewed right with a mean of 16.9 close to the median of 17 and a standard deviation of 4.3mpg. The boxplot indicates that there are at least 4 upper end outliers achieving a city MPG of approximately over 28 mpg. 

