---
title: "speech_in_class"
output: 
  html_document:
   highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Reading in my data from google doc
```{r}
speech <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vRq8oZv7Hry8oUk6ApjnXU6aCYa1vbCLwQsiaHu6uSK5w9Z4NmffV_EUdt_xGepnFzdFKBzg3Cqirej/pub?gid=0&single=true&output=csv")
```

Calculate the Average Word Length (Mean)
```{r}
mean(speech$Length)
```

Take a random sample of 10 rows from the `speech` dataframe, using `sample_n`.  The `sample_n` function is part of the `dplyr` package. First, we must install the package and then `library` the package

library the `dplyr` package
```{r, message=FALSE}
library(dplyr)
```

random sample of 10 rows from the `speech` dataframe
```{r}
speech_10_random <- sample_n(speech,10)
```

View sample
```{r}
speech_10_random

```

Calculate mean word length for my random sample
```{r}
mean(speech_10_random$Length)
```

Sample and Calculate Mean in One Chunk
```{r}
speech_10_random_run <- sample_n(speech,10)
mean(speech_10_random_run$Length)
```















