---
title: "ACS2018-Recode"
author: "Jose Rico"
date: "6/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
```

3.  Load in your data.

Load in your raw data file - with `load` command for the `addhealth4` data
```{r}
#load("/cloud/project/data/addhealth4.RData") ##RStudioCloud
ACS2018_TriCity <- readRDS("~/R/spring2020/data/ACS2018_TriCity.rds") ##RStudio Desktop
```

```{r}
tabyl(ACS2018_TriCity$COW)
```
Codebook

CIT Character 1
Citizenship status
1 .Born in the U.S.
2 .Born in Puerto Rico, Guam, the U.S. Virgin Islands, or the .Northern Marianas
3 .Born abroad of American parent(s)
4 .U.S. citizen by naturalization
5 .Not a citizen of the U.S.

COW Character 1
Class of worker
b .N/A (less than 16 years old/NILF who last worked more than 5 .years ago or never worked)
1 .Employee of a private for-profit company or business, or of an .individual, for wages, salary, or commissions
2 .Employee of a private not-for-profit, tax-exempt, or .charitable organization
3 .Local government employee (city, county, etc.)
4 .State government employee
5 .Federal government employee
6 .Self-employed in own not incorporated business, professional .practice, or farm
7 .Self-employed in own incorporated business, professional .practice or farm
8 .Working without pay in family business or farm
9 .Unemployed and last worked 5 years ago or earlier or never .worked

ENG Character 1
Ability to speak English
b .N/A (less than 5 years old/speaks only English)
1 .Very well
2 .Well
3 .Not well
4 .Not at all

FER Character 1
Gave birth to child within the past 12 months
b .N/A (less than 15 years/greater than 50 years/ male)
1 .Yes
2 .No

LANX Character 1
Language other than English spoken at home
b .N/A (less than 5 years old)
1 .Yes, speaks another language
2 .No, speaks only English

Select variables
```{r}
acs <- ACS2018_TriCity %>%  # page 30 of recode book
  select(PWGTP, SEX, AGEP, CIT, COW, ENG, FER, JWMNP, LANX, MAR, SEX, SCHL, WAGP, ESR, WKHP , WRK, NATIVITY, PINCP )
```

Recode variables
```{r}
acs_c <- acs %>% 
  mutate(ENG_c = recode_factor(ENG,
                                "1"="Very Well",
                                "2"="Well",
                                "3"="Not Well",
                                "4"="Not at all",
                                .ordered=T)) %>%
  mutate(CIT_c = recode_factor(CIT,
                                "1"="Born in US",
                                "2"="Born in US Territory",
                                "3"="Born Abroad by American Parent",
                                "4"="US Citizen by Naturalization",
                               "5"="Not a citizen of the US",
                                .ordered=T)) %>% 
   mutate(Work_Class = recode_factor(COW,   
      # N/A (less than 16 years old/NILF who last worked more than 5 .years ago or never worked)
                                "1"="Private",
                                "2"="Private",
                                "3"="Government",
                                "4"="Government",
                               "5"="Government",
                               "6"="Self-Employed",
                                "7"="Self-Employed",
                               "8"="Working for family without pay",
                               "9"="Unemployed",
                                .ordered=T)) %>% 
  mutate(FER_c = recode_factor(FER,
                                "1"="Yes",
                                "2"="No",
                                .ordered=T)) %>% 
    mutate(SEX_c = recode_factor(SEX,
                                "1"="Male",
                                "2"="Female",
                                .ordered=T)) %>% 
   mutate(AGE_C = cut(AGEP, breaks=c(-Inf,5.5,13.5,18.5,30.5,64.5,75, Inf),
                      labels=c("5 and under", 
                               "6-13",
                               "14-18",
         "19-30",
         "31-64",
         "65-75",
         "over 75" ),
          ordered_result = TRUE))
  
```

```{r}
acs_c %>% 
  tabyl(AGE_C, Work_Class) %>% 
    adorn_totals(c("row","col")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
    adorn_ns() %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left")  %>% 
    add_header_above(c("","Work Class"= 7 ))
```

```{r}
acs_c %>% 
  tabyl(SEX_c, Work_Class) %>% 
    adorn_totals(c("row","col")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
    adorn_ns() %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left")  %>% 
    add_header_above(c("","Work Class"= 7 ))
```



```{r}
acs_c %>% 

  #filter(ENG_c %in% c("Very Well", "Well", "Not Well", "Not at all")) %>% 
  #droplevels() %>% 

  
  tabyl(CIT_c, ENG_c) %>% 
    adorn_totals(c("row","col")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
    adorn_ns() %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left")  %>% 
    add_header_above(c("","Ability to Speak English"= 6 ))
```





