---
title: "hw05_explore_relationships"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    css: /cloud/project/css/sandstone.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, messages=F, warnings=F)
```

Note: you will need to run the following in your console.  ##  devtools::install_github("haleyjeppson/ggmosaic")  ##

# Set-up
```{r library packages, results='hide', message=FALSE}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
library(ggmosaic)
library(scales)
```

```{r set ggplot2 parameters, echo=F}
theme_update(plot.title = element_text(hjust = 0.5))
```

## Load data frame from HW 3 - Data Management
```{r load data}
load("/cloud/project/data/demo_addhealth_clean.Rdata")
```

# Table of Variables

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
RACE  |  race_c  | subject's race
BIO_SEX4 | gender | subject's gender
--  |  age | subject's age
H4ED2 | college_deg | Did you complete college (bachelor's degree)?
H4ED2 | educ_completed_5 | Highest level of education completed (5 levels)
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4MH23 | too_tired | (During the past seven days:) You felt that you were too tired to do things.
H4LM19 | hrs_worked_week | How many hours a week (do/did) you usually work at this job?



# Association between categorical variables

## Analysis of Relationship between Education Level and Engaging in Vigorous Activity

### Contingency (two-way) table
```{r summary table cat-cat 1}
demo %>%
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  tabyl(vigorous_activity, educ_completed_5) %>%
  adorn_totals(c("row","col")) %>% 
  adorn_percentages("col") %>% 
  adorn_pct_formatting(digits=1) %>% 
  adorn_ns() %>% 
  kable() %>%
  kable_styling(full_width = F, position="left") %>% 
  add_header_above(c("college degree"=7))
  
```


### Mosaic Plot
```{r mosiac plot cat-cat 1}
demo %>%
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  ggplot() +
    geom_mosaic(aes( x = product(vigorous_activity, educ_completed_5), fill=vigorous_activity), na.rm=TRUE)+
    ggtitle("Engaging in a Vigorous Activity (24h) comparison with Education Level ") 
```

### Narrative Summary

<div id="narrative">
The percent of those who engaging in vigorous activity in the past 24 hours is decreasing as education level increase, showing a potential association.
</div>


# Association between numerical and categorical variables

## Analysis of Relationship between Educational Level anfd Usual Hours Worked Per Week 

### Summary Statistics 
```{r summary table num-cat 1}
demo %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  group_by(educ_completed_5) %>% 
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = mean(hrs_work_week),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = sd(hrs_work_week)) %>% 
  kable() %>% 
    kable_styling(full_width = F, 
                  position="left")   %>% 
      add_header_above(c("hours worked summary statistics "= 9))
```

### Faceted Histogram

```{r}
demo %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(hrs_work_week))+
  geom_histogram(aes(y=..density..), color="white", fill="steelblue", binwidth = 5)+
    facet_wrap(~educ_completed_5, ncol=1)
```

### Boxplots
Probably a better way to compare distributions is with boxplots

```{r}
demo %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(y=hrs_work_week, x=educ_completed_5))+
  geom_boxplot(aes(fill=educ_completed_5))+  # potentially add , outlier.shape = NA
  coord_flip()  # might be nice to re-scale axis 
  
```

### Violin Plots
```{r}
demo %>% 
  filter(!educ_completed_5 =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  ggplot(aes(y=hrs_work_week, x=educ_completed_5))+
  geom_violin(aes(fill=educ_completed_5), adjust=2)+  # potentially add , outlier.shape = NA
  coord_flip()  # might be nice to re-scale axis 
  
```






