---
title: "hw05_explore_relationships_sample_report"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    #css: ~/R/spring2020/data/sandstone.css
    tables: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, messages=F, warnings=F)
```

```{r library packages, results='hide', warnings=F, message=FALSE}
library(dplyr)
library(ggplot2)
library(janitor)
library(purrr)
library(kableExtra)
library(ggmosaic)  ###Note: you will need to run the following in your console.###  ###devtools::install_github("haleyjeppson/ggmosaic")###
library(scales)
```

```{r set ggplot2 parameters, echo=F}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load data}
load("~/R/spring2020/data/demo_addhealth_clean.Rdata")
```
 

# Variables of Interest

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
BIO_SEX4 | gender | subject's gender
H4ED2 | education | Highest level of education completed (5 levels)
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4LM19 | hrs_worked_week | How many hours a week (do/did) you usually work at this job?

# Part One : Relationship between categorical variables

Subject were asked "*In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath*?  First we will break this question down by education level, then by gender.  Then we will group by gender and compare response by education level.

## Vigorous Activity Proportion by Educational Level

```{r table sweat/educ w.filter}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  tabyl( education, vigorous_activity) %>%
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left") %>% 
    add_header_above(c("","vigorous activity"= 2))
```

```{r bar graph sweat/educ}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = T), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Vigorous Activity  by Gender
```{r table sweat/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  
  tabyl(gender, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 

  kable() %>%
    kable_styling(bootstrap_options = "striped",
                full_width = T, 
                position="left") %>% 
    add_header_above(c("","vigorous activity"=2))
  
```

```{r bar graph sweat/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  
  ggplot(aes(x=gender, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = TRUE), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

# Part Two: Group by a Third Categorical Variable
## Vigorous Activity by Education Level grouped by Gender
```{r table sweat/edu/gendermale} 
demo %>% 
  filter(gender =="male") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
  
  kable() %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>%
    add_header_above(c("","MALE RESPONSES \n \n vigorous activity"=2))

```

```{r table sweat/educ/genderfemale}
demo %>% 
  
  filter(gender == "female") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 

  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>% 
    add_header_above(c("","FEMALE RESPONSES \n \n vigorous activity"=2))

```

```{r bar plot sweat/educ/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = TRUE), width=.6) +
    
    scale_fill_brewer(palette="Dark2") +
    facet_wrap(~gender, ncol=1) +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Narrative Summary

<div id="narrative">
When looking at the all `addhealth4`observations, the percent of those who engaging in vigorous activity in the past 24 hours decreases as education level increases, showing a potential association.  When comparing vigorous activity with gender, **49% of men** responsed with **yes**, compared to **33% for women**.  The relationship between vigorous activiy and education level was different for men and women. The earlier trend of vigorous activity decreasing and education level increases is observed for men, but for women the trends is not observed.
</div>



# Part Three: Relationship between numerical and categorical variables

## Analysis of Relationship between Usual Hours Worked Per Week and Educational Level

### Summary Statistics 
```{r summary table hours/educ}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  group_by(education) %>%
  
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = round(mean(hrs_work_week),1),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = round(sd(hrs_work_week),1)) %>% 
  
  kable() %>% 
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left")   %>% 
    add_header_above(c("hours worked summary statistics "= 9))
```

### Visualization (Use the best option)

```{r histogram hours/educ, eval = F}
demo %>% 
  
  filter(!education =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  
  ggplot(aes(hrs_work_week))+
  
    geom_histogram(aes(y=..density..,
                       fill=education), 
                   color="white", 
                   binwidth = 5) +
  
      scale_fill_brewer(palette="Dark2") +
  
    facet_wrap(~ education, 
               ncol=1,
               as.table = FALSE)  ## reverses the order of the levels to match boxplots
```

```{r density plot hours/educ, eval=F}
demo %>% 
  
  filter(!education =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  
  ggplot(aes(hrs_work_week)) +
  
    geom_density(aes(fill=education), 
                 color="white",
                 adjust=2) +
  
    scale_fill_brewer(palette="Dark2") +
  
    facet_wrap(~ education, 
               ncol=1,
               as.table = FALSE)  ## reverses the order of the levels to match boxplots
  
```

```{r boxplot hours/educ, eval=T}
demo %>% 
  
  filter(!education =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  
  ggplot(aes(y=hrs_work_week, x=education))+
  
    scale_fill_brewer(palette="Dark2") +
  
    geom_boxplot(aes(fill=education)) +
  
    coord_flip()  # might be nice to re-scale axis ( to see middle 50 better)
  
```

```{r violin plot hours/educ, eval=F}
demo %>% 
  
  filter(!education =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  
  ggplot(aes(y=hrs_work_week, x=education)) +
  
    geom_violin(aes(fill=education),
                draw_quantiles=c(0.25,0.5,.75),
                adjust=2) +
  
    scale_fill_brewer(palette="Dark2") +
  
    coord_flip()  
  
```

```{r density curves same plot hours/educ, eval=F}
demo %>% 
  
  filter(!education =="don't know") %>% 
  droplevels() %>% 
  filter(hrs_work_week < 121) %>% 
  
  ggplot(aes(x=hrs_work_week, color=education)) +
  
    geom_density(alpha = 0.1, adjust=2) +
  
    xlim(10, 80) +
  
    scale_fill_brewer(palette="Dark2") 
```




