---
title: "HW05 Exploring Relationships Sample Report"
author: "your name"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    css: /cloud/project/css/sandstone.css
    tables: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, messages=F, warnings=F)
```

```{r library packages, echo=F, results='hide', warnings=F, message=FALSE}
library(dplyr)
library(ggplot2)
library(janitor)
library(purrr)
library(kableExtra)
library(ggmosaic)  ###Note: you will need to run the following in your console.###  ###devtools::install_github("haleyjeppson/ggmosaic")###
library(scales)
```

```{r set ggplot2 parameters, echo=F}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load demo data, echo=F}
load("/cloud/project/data/demo_addhealth_clean.Rdata")
```
 

# Variables of Interest

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
BIO_SEX4 | gender | subject's gender
H4ED2 | education | Highest level of education completed (5 levels)
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4LM19 | hrs_worked_week | How many hours a week (do/did) you usually work at this job?

# Part One : Relationship between categorical variables

Subject were asked "*In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath*?  First we will break this question down by education level, then by gender.  Then we will group by gender and compare response by education level.

## Vigorous Activity Proportion by Educational Level

```{r table sweat/educ w.filter}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  tabyl( education, vigorous_activity) %>%
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left") %>% 
    add_header_above(c("","vigorous activity"= 2))
```

```{r bar graph sweat/educ}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = T), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Vigorous Activity  by Gender
```{r table sweat/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  
  tabyl(gender, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 

  kable() %>%
    kable_styling(bootstrap_options = "striped",
                full_width = T, 
                position="left") %>% 
    add_header_above(c("","vigorous activity"=2))
  
```

```{r bar graph sweat/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no"))  %>% 
  droplevels() %>% 
  
  ggplot(aes(x=gender, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = TRUE), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

# Part Two: Group by a Third Categorical Variable
## Vigorous Activity by Education Level grouped by Gender
```{r table sweat/edu/gendermale} 
demo %>% 
  filter(gender =="male") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
  
  kable() %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>%
    add_header_above(c("","MALE RESPONSES \n \n vigorous activity"=2))

```

```{r table sweat/educ/genderfemale}
demo %>% 
  
  filter(gender == "female") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 

  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>% 
    add_header_above(c("","FEMALE RESPONSES \n \n vigorous activity"=2))

```

```{r bar plot sweat/educ/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = TRUE), width=.6) +
    
    scale_fill_brewer(palette="Dark2") +
    facet_wrap(~gender, ncol=1) +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Narrative Summary

<div id="narrative">
When looking at the all `addhealth4`observations, the percent of those who engaging in vigorous activity in the past 24 hours decreases as education level increases, showing a potential association.  When comparing vigorous activity with gender, **49% of men** responsed with **yes**, compared to **33% for women**.  The relationship between vigorous activiy and education level was different for men and women. The earlier trend of vigorous activity decreasing and education level increases is observed for men, but for women the trends is not observed.
</div>



# Part Three: Relationship between numerical and categorical variables

## Analysis of Relationship between Usual Hours Worked Per Week and Educational Level

### Summary Statistics 
```{r summary table hours/educ}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  group_by(education) %>%
  
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = round(mean(hrs_work_week),1),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = round(sd(hrs_work_week),1)) %>% 
  
  kable() %>% 
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left")   %>% 
    add_header_above(c("hours worked summary statistics "= 9))
```

### Visualization (Use the best option)

```{r histogram hours/educ, eval = T}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(hrs_work_week))+
  
    geom_histogram(aes(y=..density..,
                       fill=education), 
                   color="white", 
                   binwidth = 5) +
  
      scale_fill_brewer(palette="Dark2") +
  
    facet_wrap(~ education, 
               ncol=1,
               as.table = FALSE)  ## reverses the order of the levels to match boxplots
```

```{r density plot hours/educ, eval=F}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(hrs_work_week)) +
  
    geom_density(aes(fill=education), 
                 color="white",
                 adjust=2) +
  
    scale_fill_brewer(palette="Dark2") +
  
    facet_wrap(~ education, 
               ncol=1,
               as.table = FALSE)  ## reverses the order of the levels to match boxplots
  
```

```{r boxplot hours/educ, eval=T}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>%  
  
  ggplot(aes(y=hrs_work_week, x=education))+
  
    scale_fill_brewer(palette="Dark2") +
  
    geom_boxplot(aes(fill=education)) +
  
    coord_flip()  # might be nice to re-scale axis ( to see middle 50 better)
  
```

```{r violin plot hours/educ, eval=F}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(y=hrs_work_week, x=education)) +
  
    geom_violin(aes(fill=education),
                draw_quantiles=c(0.25,0.5,.75),
                adjust=2) +
  
    scale_fill_brewer(palette="Dark2") +
  
    coord_flip()  
  
```

```{r density curves same plot hours/educ, eval=F}
demo %>% 
  
  filter(hrs_work_week < 121 &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(x=hrs_work_week, color=education)) +
  
    geom_density(alpha = 0.1, adjust=2) +
  
    xlim(10, 80) +
  
    scale_fill_brewer(palette="Dark2") 
```

### Narrative Summary

<div id="narrative">
When looking at the all `addhealth4`observations, the *percent of those who engaging in vigorous activity in the past 24 hours decreases as education level increases*, showing a potential association.  When comparing vigorous activity with gender, **49% of men** responsed with **yes**, compared to **33% for women**, for difference in sample proportions of **16%**.  The relationship between vigorous activiy and education level was different for men and women. The earlier trend of vigorous activity decreasing and education level increases is observed for men, but for women the trends is not observed.
</div>

# Part Four : Group by a Second Categorical Variable 
We will be adding `gender` into the examination of hours work and education level.  So, first we will examine the relationship between hours worked and gender, then we will assess how hours worked by education level changes when we group by a second categorical variable of gender.

## Hours worked broken down by Gender 

### Numerical Summary
Same commands as earlier for hours/education, except now we will `group_by(gender)`.
```{r summary table hours/gender}
demo %>% 
  
  filter(hrs_work_week < 121) %>%  
  droplevels() %>% 
  
  group_by(gender) %>%
  
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = round(mean(hrs_work_week),1),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = round(sd(hrs_work_week),1)) %>% 
  
  kable() %>% 
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left")   %>% 
    add_header_above(c("hours worked summary statistics "= 9))
```

### Visualization - Faceted Histogram

```{r histogram hours/gender}
demo %>% 
  
  filter(hrs_work_week < 121) %>% 
  droplevels() %>% 
  
  ggplot(aes(hrs_work_week))+
  
    geom_histogram(aes(y=..density..,
                       fill=gender), 
                   color="white", 
                   binwidth = 5) +
  
      scale_fill_brewer(palette="Dark2") +
  
    facet_wrap(~ gender, 
               ncol=1,
               as.table = FALSE)  ## reverses the order of the levels to match boxplots
```

## Examine Hours Worked by Gender grouped Education Level 
To create summary statistics grouped by a single variable, the process is similar to what we did earlier for vigorous activity.  Just add a filter command for each level of your categorical variable.  Here we are grouping by gender so, we are creating two summary tables for each gender.  

### Summary Tables
```{r summary table hours/educ/gendermale}
demo %>% 
  
  filter(gender == "male") %>% 
  
  filter(hrs_work_week < 121) %>%  
  droplevels() %>% 
  
  group_by(education) %>%
  
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = round(mean(hrs_work_week),1),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = round(sd(hrs_work_week),1)) %>% 
  
  kable() %>% 
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left")   %>% 
    add_header_above(c("MALE RESPONSES \n \n hours worked summary statistics "= 9))
```

```{r summary table hours/educ/genderfemale}
demo %>% 
  
  filter(gender == "female") %>% 
  
  filter(hrs_work_week < 121) %>%  
  droplevels() %>% 
  
  group_by(education) %>%
  
  summarize(n = n(),
            min = min(hrs_work_week),
            q1 = quantile(hrs_work_week, probs=0.25),
            median = median(hrs_work_week),
            mean = round(mean(hrs_work_week),1),
            q3 = quantile(hrs_work_week, probs=0.75),
            max = max(hrs_work_week),
            sd = round(sd(hrs_work_week),1)) %>% 
  
  kable() %>% 
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left")   %>% 
    add_header_above(c("FEMALE RESPONSES \n \n hours worked summary statistics "= 9))
```

### Boxplot
In order to include an additional categorical variable add the `aes`thetic `color=gender` to your `ggplot` command.  This will split each educational level by gender.
```{r boxplot hours/educ/gender}
demo %>% 
  
  filter(hrs_work_week < 121) %>% 
  droplevels() %>% 
  
  ggplot(aes(y=hrs_work_week, x=education, color=gender))+
  
    scale_fill_brewer(palette="Dark2") +
  
    geom_boxplot() +
  
    coord_flip()  # might be nice to re-scale axis ( to see middle 50 better)
  
```

### Narrative Summary
<div id="narrative">
When we look closer at usual hours worked by education level, but now grouped by gender - there are some interesting observations.  For each level of education the mean hours for women is less than that of men.  The difference in the range of middle 50% can be seen in the boxplot above.  Only when we examine **adv degree** do we see the middle 50% observations align with eachother across gender.
</div>
