---
title: "Week Three Assignment - Data Preparation"
author: "your name here"
output:
  html_document:
    theme: sandstone
    highlight: tango
    toc: yes
    toc_float: yes
    css: ../css/sandstone.css
  pdf_document: default
---


```{r, echo=F}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE) #Chunk default settings 
```

## Project Set-up
Library necessary packages 
```{r}
library(dplyr)
library(janitor)
```

## Load in original data 
Load in your raw data file - with `load` command for the `addhealth4` data
```{r}
#load("/cloud/project/data/addhealth4.RData")
```


## New data frame with selected variables 
```{r}
#mydata <- addhealth4 %>% 
  #select(RACE, BIO_SEX4, iyear4, h4od1y , H4GH1, H4GH7, H4GH8, H4MH2, H4MH7, H4MH9, H4MH22, H4MH23, H4MH24, H4MH25, H4DA11 )
```


## Recoding of variables

### Variable 1

### Variable 2

### Variable 3

### Variable 4

## Cleaned data frame

**Step 1** : Select only the variables that you want to keep, so you only have the version of the variables that you want in your data frame.
```{r}
#clean <- mydata_c %>% select(race_c, BIO_SEX4, iyear4, h4od1y , H4GH1, H4GH7, H4GH8, H4MH2, H4MH7, H4MH9, H4MH22, H4MH23, H4MH24, H4MH25, vigorous_activity)
```

Alternative Approach to Step 1: You can also use `select` to remove variable with the `-variable`
```{r}
#clean <- mydata_c %>% select(-RACE, -H4DA11)
```

**Step 2**: Save the resulting data set to your `data` folder as `datasetname_clean.Rdata` e.g. `addhealth_clean.Rdata`. 
```{r}
#save(clean, file="/cloud/project/data/myaddhealth_clean.Rdata")
```



## Table of variables

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
addhealthvar  |  my_addhealthvar  | description (from addhealth)


