---
title: "HW 7 - Single Variable Analysis w/ Inference - Guide"
author: "Jose Rico"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float : true
    css: /cloud/project/css/sandstone.css
    theme: sandstone
    highlight: tango
---

```{r set chunk option, echo=F}
knitr::opts_chunk$set(echo = F, warning=FALSE, message=FALSE, fig.height=4, fig.align='center') #Chunk default settings 
```

```{r library packages}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
library(mosaic)
```

```{r set ggplot2 parameters}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load data}
load("/cloud/project/data/demo_addhealth_clean.Rdata")

```


# Table of Variables

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
RACE  |  race_c  | subject's race
BIO_SEX4 | gender | subject's gender
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?
H4MH23 | too_tired | (During the past seven days:) You felt that you were too tired to do things.

# 1st Categorical (binary - 2 levels) `vigorous_activity`

**Numerical Summary Table**  

```{r summary table cat 1}
  demo %>% 
  filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  droplevels() %>% 
  tabyl(vigorous_activity) %>%
  adorn_pct_formatting(digits=1) %>%
  adorn_totals()%>%
  kable() %>%
  kable_styling(full_width = F, position="center")
```

**Bar Plot - Frequency of Levels**

```{r bar plot cat 1}
demo %>% 
  filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  droplevels() %>% 
  ggplot(aes(x=vigorous_activity, y= ..prop.., group=1)) +
    geom_bar(fill="steelblue", width=0.5) +
      ggtitle("In the past 24 hours, have you participated in vigorous physical activity long 
enough to work up a sweat, get your heart thumping, or get out of breath?") +
      xlab("subject's response")
  
```

**Inference (CI and Test of Significance)**
```{r}
##clean variable for inference tests - binary (only 2 levels)
vig_act <- demo %>% 
  select(vigorous_activity) %>% 
  filter(vigorous_activity=="yes" | vigorous_activity=="no") %>% 
  droplevels() 
```

```{r, echo=T}
## inference CI and test of significance  (note: prop.test is for binary variables)
prop.test(~vigorous_activity == "no", data = vig_act)
```

```{r, echo=F, eval=F}
## the xchisq.test assesses the probability of obtaining a sample result if the levels were equaly likely.
## chisquared not needed for binart categorical variables
xchisq.test(~vigorous_activity, data = vig_act)
```

**Summary of Analysis**

<div id = "narrative">
The participants in the **addhealth** study were asked, "*In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?*  

* 3056 out of 5111 (59.8%) of participants in the study answered "no".  
* Our sample proportion of 59.8% represents a statistically significant majority ($h_{0}=0.5$, p-value < 2.2e-16)
* 95% confidence interval for non-participation in vigorous activity is 58.4% to 61.1%  (output from **mosaic**'s `prop.test` function)
</div>

# 2nd Categorical (mulitple levels) `too tired`

**Numerical Summary Table**  

```{r}
  demo %>% 
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time")) %>% 
  droplevels() %>% 
  tabyl(too_tired) %>%
  adorn_pct_formatting(digits=1) %>%
  adorn_totals()%>%
  kable() %>%
  kable_styling(full_width = F, position="center")
```

**Bar Plot - Frequency of Levels**

```{r}
demo %>% 
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time")) %>% 
  droplevels() %>% 
  ggplot(aes(x=too_tired, y= ..prop.., group=1)) +
    geom_bar(fill="steelblue", width=0.5) +
      ggtitle("During the past seven days: \nYou felt that you were too tired to do things") +
      xlab("subject's response")
  
```

**Inference (CI and Test of Significance)**
```{r}
##clean variable for inference tests - binary (only 2 levels)
tired <- demo %>% 
  select(too_tired) %>% 
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time")) %>% 
  droplevels() 
```

```{r, echo=T}
## inference CI and test of significance                    ##
## (note: prop.test needs you to choose a particular level) ##
prop.test(~too_tired == "never or rarely", data = tired)
```

```{r, echo=T}
## the xchisq.test assesses the probability of obtaining ##
## a sample result if the levels were equaly likely.     ##
xchisq.test(~too_tired, data = tired)  ## you can add % for each level with p(c(.25,.25,.25,.25))
```

**Summary of Analysis**

<div id = "narrative">
The participants in the **addhealth** study were asked, "*During the past seven days: You felt that you were too tired to do things*"  

* 1764 out of 5113 (**34.5%**) of participants in the study answered **"never or rarely"**.  
* Our sample proportion of **34.5%** represents a **statistically significant minority**  ($h_{0}=0.5$, p-value < 2.2e-16)
* 95% confidence interval for **never or rarely felt too tired to do thing** is **33.2% to 35.8%**  (output from **mosaic**'s `prop.test` function)
* A Chi-Squared test for equal likely levels yielded a test statistics of $\chi^2=2560$, with $p-value < 2.2e-16$, which allows us to conclude that the proportion for each response level are statistically significant (not equally likely).  *In other words our sample's distribution of results is not likely to occur by chance, if each response was equally likely*
</div>




