---
title: "HW08 Exploring Relationships w/ Inference - Guide"
author: "Jose Rico"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float : true
    css: /cloud/project/css/sandstone.css
    theme: sandstone
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, messages=F, warnings=F)
```

```{r library packages, echo=F, results='hide', warnings=F, message=FALSE}
library(dplyr)
library(ggplot2)
library(janitor)
library(kableExtra)
library(scales)
library(mosaic)
```

```{r set ggplot2 parameters, echo=F}
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r load demo data, echo=F}
load("/cloud/project/data/demo_addhealth_clean.Rdata")
```
 

# Variables of Interest

Original Variable  |   New Variable  |    Description
-------------------|-----------------|------------------
BIO_SEX4 | gender | subject's gender
H4MH23 | too_tired | (During the past seven days:) You felt that you were too tired to do things.
H4ED2 | education | Highest level of education completed (5 levels)
H4DA11 | vigorous_activity | In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath?



# Part One : Relationship between 1st set of categorical variables (`vigorous_activity` vs. `education`)

Subject were asked "*In the past 24 hours, have you participated in vigorous physical activity long enough to work up a sweat, get your heart thumping, or get out of breath*?  

## Table and Graphic of `vigorous_activity` vs. `education`

```{r}
demo %>%
  
  filter(vigorous_activity %in% c("no", "yes") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  tabyl( education, vigorous_activity) %>%
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>%
    adorn_ns() %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left") %>% 
    add_header_above(c("","vigorous activity"= 2))
```

```{r }
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = T), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Inference: Chi Squared Test for Independence 
```{r}
## filtering and storing in new dataframe
act_ed <- demo %>%
  select(education, vigorous_activity) %>% 
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() 
```

The order of the variables in the `xchisq.test` does not matter in the calculation
of the test statistic.  They are written this way so the test output matches the earlier table.
```{r, echo=T}
xchisq.test(education ~ vigorous_activity , data = act_ed) 
```

## Inference: Assessing the largest difference in proportions
From the table and Chi Squared results, the largest difference is between education levels is **hs grad** and **adv degree**.  I want to construct a confidence interval to see how big is the difference.  So, I will just filter the **hs grad** and **adv degree** levels into a new data frame and then run a `prop.test` for difference in proportions for **no** response to vigorous activity.
```{r}
act_ed2 <- demo %>%
  select(education, vigorous_activity) %>% 
  filter(vigorous_activity %in% c("no","yes") &
         education %in% c( "hs grad", "adv degree")) %>% 
  droplevels()
```

```{r, echo=TRUE}
prop.test((vigorous_activity == "no") ~ education, data = act_ed2)
```

## Narrative Summary

<div id="narrative">
NEED TO WORK ON ALONG WITH DIRECTIONS
</div>

# Part Two : Relationship between 2nd set of categorical variables (`too_tired` vs. `gender`)

Subject were asked "*During the past seven days: You felt that you were too tired to do things*"?  First we will break this question down by gender.  Then we will...

## Table and Graphic of `too_tired` vs. `gender`

```{r }
demo %>%
  
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time"),
         gender %in% c("male","female")) %>% 
  droplevels() %>% 

  tabyl(gender, too_tired) %>%
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>%
    adorn_ns() %>% 
  
  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped", full_width = T, 
                position="left") %>% 
    add_header_above(c("","How often tired in past 7 days?"= 4))
```

```{r }
demo %>%
  
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time"),
         gender %in% c("male","female")) %>% 
  droplevels() %>% 
  
  ggplot(aes(x=gender, fill=too_tired)) +
    geom_bar(position = position_fill(reverse = T), width=.6) +
  
    scale_fill_brewer(palette="Dark2") +
    scale_y_continuous(labels=percent) +
    coord_flip()
```



## Inference: Chi Squared Test for Association
```{r}
tired_gen <- demo %>%
  select(gender, too_tired) %>% 
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time"),
         gender %in% c("male","female")) %>% 
  droplevels()  
```

```{r, echo=T}
xchisq.test( gender ~ too_tired , data = tired_gen) 
```

## Inference: Assessing the largest difference in proportions
Note: this time around I am not filtering for the 2 levels where the largest difference was found, 
because my explanatory variable only has 2 levels.  Instead, I will choose the level of the largest 
difference for a single level in my response variable (**never or rarely**) only when I select the
level in `prop.test`'s argument.  (actually - I did not need to use the chunk below) 
```{r}
### unecessary chunk - since not filtering my explanatory variable
tired_gen2 <- demo %>%
  select(gender, too_tired) %>% 
  filter(too_tired %in% c("never or rarely","sometimes","alot of the time", "most of the time or all of the time"),
         gender %in% c("male","female")) %>% 
  droplevels()
```

```{r, echo=TRUE}
prop.test((too_tired=="never or rarely") ~ gender, data = tired_gen2)  ## prop.test(response ~ explanatory)
```

## Narrative Summary

<div id="narrative">
NEED TO WORK ON ALONG WITH DIRECTIONS
</div>

# Part Three: Group by a Third Categorical Variable - `vigorous_activity` vs. `education` vs. `gender`

## Graphic of Breakdown by Gender
```{r bar plot sweat/educ/gender}
demo %>%
  
  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 

  ggplot(aes(x=education, fill=vigorous_activity)) +
    geom_bar(position = position_fill(reverse = TRUE), width=.6) +
    
    scale_fill_brewer(palette="Dark2") +
    facet_wrap(~gender, ncol=1) +
    scale_y_continuous(labels=percent) +
    coord_flip()
```

## Table for Males
```{r table sweat/edu/gendermale} 
demo %>% 
  filter(gender =="male") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>%
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 
  
  kable() %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>%
    add_header_above(c("","MALE RESPONSES \n \n vigorous activity"=2))

```

## Inference: Chi Squared Test for Association (for Males only)
```{r}
act_ed_M <- demo %>%
  select(education, vigorous_activity, gender) %>% 
  filter(gender=="male",
         vigorous_activity %in% c("yes","no"),
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() 
```

```{r, echo=T}
xchisq.test( education ~ vigorous_activity , data = act_ed_M) 
```

## Table for Women
```{r table sweat/educ/genderfemale}
demo %>% 
  
  filter(gender == "female") %>% 

  filter(vigorous_activity %in% c("yes","no") &
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() %>% 
  
  tabyl(education, vigorous_activity) %>% 
    adorn_totals(c("row")) %>% 
    adorn_percentages("row") %>% 
    adorn_pct_formatting(digits=0) %>% 

  kable(booktabs = TRUE) %>%
    kable_styling(bootstrap_options = "striped",
                  full_width = T, 
                  position="left") %>% 
    add_header_above(c("","FEMALE RESPONSES \n \n vigorous activity"=2))

```

## Inference: Chi Squared Test for Association (for Females only)
```{r}
act_ed_F <- demo %>%
  select(education, vigorous_activity, gender) %>% 
  filter(gender=="female",
         vigorous_activity %in% c("yes","no"),
         education %in% c("not hs grad", "hs grad", "post hs", "bachelors", "adv degree")) %>% 
  droplevels() 
```

```{r, echo=T}
xchisq.test( education ~ vigorous_activity , data = act_ed_F) 
```



## Narrative Summary

<div id="narrative">
NEED TO WORK ON ALONG WITH DIRECTIONS
</div>


